---
title: "06-03-19 Saliva Demog Analysis"
author: "Lauren Middleton"
date: "6/3/2019"
output: html_document
---
```{r}
# Purpose: Do statistical analysis on demog_ex. Checking for differences in variables between Male and Female groups
# 
# Inputs: 06-03-19 demog_ex.csv   - 22 participants and 26 variables, file was constructed based on the results of the
#                                   survey given to the parents of participants
#         05-28-19 demog_full.csv - demographics by epithelial, immune, oragene, and whole saliva samples
#
# Outputs: 06-03-19 demog_ex_dropped.csv - dropped one participant due to their fractions not passing QC
#          stats analysis
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("ggpubr")
```

#set up;
```{r}
setwd("C:/Users/HP/Documents/Research 2019")

library(ggplot2)
library(reshape2)
library(readr)
library(RColorBrewer)
library(compareGroups)
```

#import dataset and save it as demog_ex;
```{r}
#Load demog_ex 
demog_ex <- read.csv("C:/Users/HP/Documents/Research 2019/06-03-19 demog_ex.csv",
                  header=TRUE,
                  sep="," )
#Load demog_full
demog_full <- read.csv("C:/Users/HP/Documents/Research 2019/05-28-19 demog_full.csv")

# Exclude participant ID = 8 because we didn't use any of their samples;
#keep all rows except 8;
x <- demog_ex[-c(8), ]
#visually check that ID = 8 was removed;
demog_ex <- x
rm(x)

#remove the demog_full excluded participant
x <- demog_full[-c(75),]

#Save demog_ex with n=21, not n=22
write.csv(demog_ex, file = "06-03-19 demog_ex_dropped.csv")

#Use this file to run the whole code:
demog_ex <- read.csv("C:/Users/HP/Documents/Research 2019/06-03-19 demog_ex_dropped.csv",
                     header = TRUE,
                     sep = ",")
View(demog_ex)
```


# Calculate basic stats
```{r}
#Race (0=white for white_bin)
table(demog_ex$white_bin) #10 white, 11 non-white
#(1=biracial for race_birac)
table(demog_ex$race_birac) #10 biracial, 11 one race

#Age
mean(demog_ex$age.r)
sd(demog_ex$age.r)
summary(demog_ex$age.r)

#cell count
median(demog_ex$cell_count)
sd(demog_ex$cell_count)
summary(demog_ex$cell_count)

#Viability
median(demog_ex$viability)
sd(demog_ex$viability)
summary(demog_ex$viability)

#Volume
median(demog_ex$volume)
sd(demog_ex$volume)
summary(demog_ex$volume)
```

#Add Linear regression and stats info to volume/cell count plot;
##Not really a good test b/c cell count and volume are not normal;
```{r}
#create a function to extract stats
ggplotRegression <- function (fit) {

#loads package
require(ggplot2)
#plot the scatterplot
ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  geom_point() +
  #create the regression line
  stat_smooth(method = "lm", col = "red") +
  labs(title = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                     "Intercept =",signif(fit$coef[[1]],5 ),
                     " Slope =",signif(fit$coef[[2]], 5),
                     " P =",signif(summary(fit)$coef[2,4], 5)))
}

#use the function
fit1 <- lm(demog_ex$cell_count ~ demog_ex$volume, data = demog_ex)
ggplotRegression(fit1)
```


# Wilcoxon signed-rank test for differences between continuous vars of male vs. female;
##unpaired, independant samples, categorical vs continuous;
##these are the same as are in the graphs;
```{r}
#testing wilcox for age.r
wilcox.test(demog_ex$age.r ~ demog_ex$Male, alternative = "two.sided")
#p = 0.733

for (i in c(12:18, 22:26)) {
  p_value <- wilcox.test(demog_ex[ ,i] ~demog_ex$Male, alternative = "two.sided")
  print(colnames(demog_ex[i]))
  print(p_value)
}
#confirmed age.r match
```


# Fisher exact test for categorical variables (white_bin and Sick);
```{r}
#white_bin
fisher.test(x = demog_ex$Male, y = demog_ex$white_bin, alternative = "two.sided", conf.level = 0.95)

#Sick
fisher.test(x = demog_ex$Male, y = demog_ex$Sick, alternative = "two.sided", conf.level = 0.95)
```


#testing for correlation btwn volume cell count;
##spearman is a nonparametric test between 2 continous vars;
```{r}
cor.test(demog_ex$cell_count, demog_ex$volume, method = "spearman")
```


#trying to use compareGroups - table 1 information
```{r}
#but first make labels for binary variables
#check original data
table(demog_ex$white_bin)
#relabel the white_bin
race_label <- factor(demog_ex$white_bin,
                     levels = c(0, 1),
                     labels = c("White", "POC"))
#check that the labeling went okay
table(demog_ex$white_bin, race_label)

#add race_label to demog_ex
demog_ex$race_label <- race_label

```



```{r}
#save table as "res"
#compare by Male_label
#variables to look at are listed
#2 is doing a kruskall-wallis test - this is better for nonparametric ANOVA (3 samples);
res <- compareGroups(Male_label ~ Sick
              + cell_count
              + viability
              + volume
              + age.r
              + race_label
              + cells_mL,
              data = demog_ex,
              #3 = categorical, 2 = non-normal;
              method = c(race_label = 3,
                         Sick = 3,
                         cell_count = 2,
                         viability = 2, 
                         volume = 2,
                         age.r = 2,
                         cells_mL = 2),
              alpha = 0.05,
              #true means using labels. I don't have any, but I can fix that;
              include.label = TRUE,
              Q1 = 0,
              Q3 = 1)
```

```{r}
#prints simplified table with p-values;
res
#prints detailed N, mean, sd, lcl, hcl, p-value, OR;
summary(res)
```

```{r}
#create descriptive table called restab from res;
restab <- createTable(res)
print(restab, which.table = "descr")
restab$descr #unformatted table;
# restab$descr[1,3] <- .999
```

# Manually running kruskall wallis test and compare p-values to table;
```{r}
#Kruskal-Wallis test is a non-parametric alternative version of a one-way ANOVA test
#Similar to two-sample Wilcoxon test, but can include more than two groups;
#age;
kruskal.test(demog_ex$age.r ~ demog_ex$Male_label, data = demog_ex)

#viability
kruskal.test(demog_ex$viability ~ demog_ex$Male_label, data = demog_ex)
```

# Does compareGroups run mean or median;
```{r}
#subset Female;
female_set <- subset(demog_ex, demog_ex$Male == 0, )

#age
mean(female_set$age.r) #12.14
median(female_set$age.r) #11.7

#it runs median;
```


## 7/16/19
# Check for correlations between demog variables Male vs ____
```{r}
##Wilcoxon: nonparametric, continuous var vs categorical t-test (Ho = two populations are equal)
##Chi Squared: nonparametric, categorical predictors (Ho: the populations are independent)
  ##Yates correction: "current use is limited" - Wikipedia, therefore correct = FALSE
  ##tests frequency of each category... - binary doesn't go well

#Male vs sick (Chi-sq)
#first find the counts of each option: Male/sick, male/not sick, female/sick, female/not sick
library(plyr)
freq_male_sick <- count(demog_ex, vars=c("Male", "Sick"))
#Chi-square test
male_sick <- chisq.test(freq_male_sick, correct = FALSE) #p-value = 0.1003

#Male vs cell_count
male_cell_count <- wilcox.test(cell_count ~ Male, data = demog_ex) #p-value = 0.6768

#Male vs viability
male_viability <- wilcox.test(viability ~ Male, data = demog_ex) #p-value = 0.1729

#Male vs volume
male_volume <- wilcox.test(volume ~ Male, data = demog_ex) #p-value = 0.7839

#Male vs overnight
male_overnight <- wilcox.test(overnight ~ Male, data = demog_ex) #p-value = 0.6733

#Male vs batch
freq_male_batch <- count(demog_ex, vars=c("Male", "batch"))
male_batch <- chisq.test(freq_male_batch, correct = FALSE) #p-value = 0.9794

#Male vs age (Wilcoxon)
male_age <- wilcox.test(age.r ~ Male, data = demog_ex) #p-value = 0.7333

#Male vs white_bin
freq_male_race <- count(demog_ex, vars=c("Male", "white_bin"))
male_white <- chisq.test(freq_male_race, correct = FALSE) #p-value = 0.764

#Male vs cells_mL
male_cells_mL <- wilcox.test(cells_mL ~ Male, data = demog_ex) #p-value = 0.7333
```

# Check for correlations between demog variables Cell Type vs ____
```{r}
##Wilcoxon: nonparametric, continuous var vs categorical t-test (Ho = two populations are equal)
##Chi Squared: nonparametric, categorical predictors (Ho: the populations are independent)
  ##Yates correction: "current use is limited" - Wikipedia, therefore correct = FALSE
  ##tests frequency of each category... - binary doesn't go well

#Male vs sick (Chi-sq)
#first find the counts of each option: Male/sick, male/not sick, female/sick, female/not sick
library(plyr)
freq_male_sick <- count(demog_ex, vars=c("Male", "Sick"))
View(freq_male_sick)
#Chi-square test
male_sick <- chisq.test(freq_male_sick, correct = FALSE) #p-value = 0.1003

#Male vs cell_count
male_cell_count <- wilcox.test(cell_count ~ Male, data = demog_ex) #p-value = 0.6768

#Male vs viability
male_viability <- wilcox.test(viability ~ Male, data = demog_ex) #p-value = 0.1729

#Male vs volume
male_volume <- wilcox.test(volume ~ Male, data = demog_ex) #p-value = 0.7839

#Male vs overnight
male_overnight <- wilcox.test(overnight ~ Male, data = demog_ex) #p-value = 0.6733

#Male vs batch
freq_male_batch <- count(demog_ex, vars=c("Male", "batch"))
male_batch <- chisq.test(freq_male_batch, correct = FALSE) #p-value = 0.9794

#Male vs age (Wilcoxon)
male_age <- wilcox.test(age.r ~ Male, data = demog_ex) #p-value = 0.7333

#Male vs white_bin
freq_male_race <- count(demog_ex, vars=c("Male", "white_bin"))
male_white <- chisq.test(freq_male_race, correct = FALSE) #p-value = 0.764

#Male vs cells_mL
male_cells_mL <- wilcox.test(cells_mL ~ Male, data = demog_ex) #p-value = 0.7333
```


# Check for correlations between demog variables Sick vs ____
```{r}
#Sick vs cell count
sick_cellcount <- wilcox.test(cell_count ~ Sick, data = demog_ex) #p-value = 0.7398

#Sick vs viability
sick_viability <- wilcox.test(viability ~ Sick, data = demog_ex) #p-value = 0.1912

#Sick vs volume
sick_volume <- wilcox.test(volume ~ Sick, data = demog_ex) #p-value = 0.8003

#Sick vs overnight
freq_sick_overnight <- count(demog_ex, vars=c("Sick", "overnight"))
sick_overnight <- chisq.test(freq_sick_overnight, correct = FALSE) #p-value = 0.09726

#Sick v batch
freq_sick_batch <- count(demog_ex, vars=c("Sick", "batch"))
sick_batch <- chisq.test(freq_sick_batch, correct = FALSE) #p-value = 0.09726

#Sick vs age (Wilcoxon)
sick_age <- wilcox.test(age.r ~ Sick, data = demog_ex) #p-value = 0.5338

#Sick vs white_bin
freq_sick_race <- count(demog_ex, vars=c("Sick", "white_bin"))
sick_white <- chisq.test(freq_sick_race, correct = FALSE) #p-value = 0.08941

#Sick vs cells_mL
sick_cells_mL <- wilcox.test(cells_mL ~ Sick, data = demog_ex) #p-value = 0.4707
```

# Check for correlations between demog variables Cell count vs ____
```{r}
#Cell count vs viability
cell_count_viability <- cor.test(demog_ex$cell_count, demog_ex$viability, method = "spearman")
cell_count_viability #p-value = 0.1764

#Cell count vs volume
cell_count_volume <- cor.test(demog_ex$cell_count, demog_ex$volume, method = "spearman")
cell_count_volume #p-value = 0.9776

#Cell count vs overnight
cell_count_overnight <- wilcox.test(cell_count ~ overnight, data = demog_ex) #p-value = 0.5476

#Cell count vs age
cell_count_age <- cor.test(demog_ex$cell_count, demog_ex$age.r, method = "spearman")
cell_count_age #p-value = 0.2719

#Cell count vs batch
cell_count_batch <- cor.test(demog_ex$cell_count, demog_ex$batch, method = "spearman")

#Cell count vs white_bin
cell_count_white <- wilcox.test(cell_count ~ white_bin, data = demog_ex) #p-value = 0.09862

#****************
#Cell count vs cells_mL 
cell_count_mL <- cor.test(demog_ex$cell_count, demog_ex$cells_mL, method = "spearman")
cell_count_mL #p-value = 4.527e-06
#****************
```

# Check for correlations between demog variables Viability vs ____
```{r}
#viability vs volume #LOTS OF "EXACT TIES" IN THIS COMPARISON
viability_volume <- cor.test(demog_ex$viability, demog_ex$volume, method = "spearman")
viability_volume #p-value = 0.2028

#viability vs overnight
viability_overnight <- wilcox.test(viability ~ overnight, data = demog_ex) #p-value = 0.1265

#viability vs age
viability_age <- cor.test(demog_ex$viability, demog_ex$age.r, method = "spearman")
viability_age #p-value = 0.4201

#viability vs batch
viability_batch <- kruskal.test(demog_ex$viability ~ demog_ex$batch, data = demog_ex)

#viability vs white_bin
viability_white <- wilcox.test(viability ~ white_bin, data = demog_ex) #p-value = 0.4179

#viability vs cells_mL
viability_mL <- cor.test(demog_ex$viability, demog_ex$cells_mL, method = "spearman")
viability_mL #p-value = 0.3404
```

# Check for correlations between demog variables Volume vs ____
```{r}
#volume vs overnight
volume_overnight <- wilcox.test(volume ~ overnight, data = demog_ex) #p-value = 0.3606

#volume vs age
volume_age <- cor.test(demog_ex$volume, demog_ex$age.r, method = "spearman")
volume_age #p-value = 0.3553

#volume vs white_bin
volume_white <- wilcox.test(volume ~ white_bin, data = demog_ex) #p-value = 0.08892

#volume vs batch
volume_batch <- kruskal.test(demog_ex$volume ~ demog_ex$batch, data = demog_ex)

#volume vs cells_mL
volume_mL <- cor.test(demog_ex$volume, demog_ex$cells_mL, method = "spearman")
volume_mL #p-value = 0.1534
```

# Check for correlations between demog variables Overnight vs ____
```{r}
#******************
#overnight vs age
overnight_age <- wilcox.test(age.r ~ overnight, data = demog_ex) #p-value = 0.00285
#******************

#overnigth vs batch
freq_overnight_batch <- count(demog_ex, vars=c("overnight", "batch"))
View(freq_overnight_batch)
overnight_batch <- chisq.test(freq_overnight_batch, correct = FALSE) #p-value = 0.6104

#overnight vs white_bin
freq_overnight_race <- count(demog_ex, vars=c("overnight", "white_bin"))
View(freq_overnight_race)
overnight_white <- chisq.test(freq_overnight_race, correct = FALSE) #p-value = 0.2962

#overnight vs cells_mL
overnight_mL <- wilcox.test(cells_mL ~ overnight, data = demog_ex) #p-value = 0.3127
```

# Check for correlations between demog variables batch vs ____
```{r}
#batch vs age
batch_age <- kruskal.test(demog_ex$age.r ~ demog_ex$batch, data = demog_ex)

#batch vs white_bin
freq_batch_race <- count(demog_ex, vars=c("batch", "white_bin"))
View(freq_overnight_race)
batch_white <- chisq.test(freq_overnight_race, correct = FALSE) #p-value = 0.2962

#batch vs cells_mL
batch_mL <- kruskal.test(demog_ex$cells_mL ~ demog_ex$batch, data = demog_ex)

#batch vs overnight
freq_overnight_batch <- count(demog_ex, vars=c("overnight", "batch"))
View(freq_overnight_batch)
overnight_batch <- chisq.test(freq_overnight_batch, correct = FALSE) #p-value = 0.6104
```

# Check for correlations between demog variables Age vs ____, white_bin vs cells/mL
```{r}
#age vs white_bin
age_white <- wilcox.test(age.r ~ white_bin, data = demog_ex) #p-value = 0.9177

#age vs cells_mL
age_mL <- cor.test(demog_ex$age.r, demog_ex$cells_mL, method = "spearman")
age_mL #p-value = 0.1688

#white_bin vs cells_mL
white_mL <- wilcox.test(cells_mL ~ white_bin, data = demog_ex) #p-value = 0.3494
```

# Set up the matrix for the heatmap
```{r}
#first generate the matrix
p.mat<-matrix(nrow=10, ncol=10)
#name the rows of the matrix
rownames(p.mat)<-c("Male",
                   "Sick",
                   "cell_count",
                   "viability",
                   "volume",
                   "overnight",
                   "batch",
                   "age.r",
                   "white_bin",
                   "cells_mL")
#name the columns the same as the rows of the matrix
colnames(p.mat)<-rownames(p.mat)
```


# Save all the p-value results into the matrix
```{r}
#male_sick has the results of all the previous tests
str(male_sick)
p.mat['Male','Sick'] <- male_sick$p.value
p.mat["Male", "cell_count"] <- male_cell_count$p.value
p.mat["Male", "viability"] <- male_viability$p.value
p.mat["Male", "volume"] <- male_volume$p.value
p.mat["Male", "overnight"] <- male_overnight$p.value
p.mat["Male", "batch"] <- male_batch$p.value
p.mat["Male", "age.r"] <- male_age$p.value
p.mat["Male", "white_bin"] <- male_white$p.value
p.mat["Male", "cells_mL"] <- male_cells_mL$p.value

#Sick
p.mat["Sick", "cell_count"] <- sick_cellcount$p.value
p.mat["Sick", "viability"] <- sick_viability$p.value
p.mat["Sick", "volume"] <- sick_volume$p.value
p.mat["Sick", "overnight"] <- sick_overnight$p.value
p.mat["Sick", "batch"] <- sick_batch$p.value
p.mat["Sick", "age.r"] <- sick_age$p.value
p.mat["Sick", "white_bin"] <- sick_white$p.value
p.mat["Sick", "cells_mL"] <- sick_cells_mL$p.value

#Cell count
p.mat["cell_count", "viability"] <- cell_count_viability$p.value
p.mat["cell_count", "volume"] <- cell_count_volume$p.value
p.mat["cell_count", "overnight"] <- cell_count_overnight$p.value
p.mat["cell_count", "batch"] <- cell_count_batch$p.value
p.mat["cell_count", "age.r"] <- cell_count_age$p.value
p.mat["cell_count", "white_bin"] <- cell_count_white$p.value
p.mat["cell_count", "cells_mL"] <- cell_count_mL$p.value

#Viability
p.mat["viability", "volume"] <- viability_volume$p.value
p.mat["viability", "overnight"] <- viability_overnight$p.value
p.mat["viability", "batch"] <- viability_batch$p.value
p.mat["viability", "age.r"] <- viability_age$p.value
p.mat["viability", "white_bin"] <- viability_white$p.value
p.mat["viability", "cells_mL"] <- viability_mL$p.value

#Volume
p.mat["volume", "overnight"] <- volume_overnight$p.value
p.mat["volume", "batch"] <- volume_batch$p.value
p.mat["volume", "age.r"] <- volume_age$p.value
p.mat["volume", "white_bin"] <- volume_white$p.value
p.mat["volume", "cells_mL"] <- volume_mL$p.value

#Overnight
p.mat["overnight", "batch"] <- overnight_batch$p.value
p.mat["overnight", "age.r"] <- overnight_age$p.value
p.mat["overnight", "white_bin"] <- overnight_white$p.value
p.mat["overnight", "cells_mL"] <- overnight_mL$p.value

#Batch
p.mat["batch", "age.r"] <- batch_age$p.value
p.mat["batch", "white_bin"] <- batch_white$p.value
p.mat["batch", "cells_mL"] <- batch_mL$p.value

#Age
p.mat["age.r", "white_bin"] <- age_white$p.value
p.mat["age.r", "cells_mL"] <- age_mL$p.value

#Race
p.mat["white_bin", "cells_mL"] <- white_mL$p.value

#heatmap does not like NAs so having a p-value of 1 is pretty much an NA
p.mat[is.na(p.mat)] <- 1

View(p.mat)
```


# Heatmap of p values
```{r}
library(gplots)
my_palette <- colorRampPalette(c("red", "yellow", "white"))
heatmap.2(p.mat,
          trace = "none",
          dendrogram = "none",
          Rowv = FALSE, #how should the rows be ordered - leave the order alone
          Colv = FALSE, #how should the columns be ordered - leave the order alone
          key = TRUE, keysize = 1.5,
          col = my_palette,
          breaks = c(seq(0,0.1,length=100),
                         seq(0.11,0.2,length=100),
                         seq(0.21,1, length=100)),
          margins = c(7, 7))



```
