---
title: "06-17-19 Saliva PCA plot"
author: "Lauren Middleton"
date: "6/17/2019"
output: html_document
---
```{r}
# Purpose: Create a PCA that plots all cell types without whole or oragene
# 
# Inputs:   pd_final   - details of sample fractions (epithelial and immune)
#           beta_final - beta matrix of sample fractions (epithelial and immune)
#
# Outputs:  density plot figure
```


# Install ggfortify to make PCA
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="C:/Users/HP/Documents/Research 2019")
knitr::opts_chunk$set(root.dir="C:/Users/HP/Documents/Research 2019")

#install.packages("ggfortify")
#install.packages("grDevices")
#install.packages("DescTools")
#install.packages("shape")
#install.packages("tidyverse")
```

# Drop cd45- from beta and pd (this has already been done)
```{r}
beta_final <- readRDS("06-11-19 beta_final.rds")
pd_final <- readRDS("06-11-19 pd_final.rds")

head(pd_final)
cd45neg_samps <- pd_final$celltype == "CD45neg"
cd45neg_samps

pd_final1 <-  pd_final[!cd45neg_samps, ]
table(pd_final1$celltype)
pd_final <- pd_final1
saveRDS(pd_final, "06-11-19 pd_final.rds")

#drop from beta
beta_final1 <- beta_final[, pd_final$meth_id]
identical(pd_final$meth_id, colnames(beta_final1))
beta_final <- beta_final1
rm(beta_final1)

saveRDS(beta_final, "06-11-19 beta_final.rds")
```

# Set-up
```{r}
#Open this file for Environment
load("06-17-19 Saliva PCA plot Environment.RData")

#load beta_final;
beta_final <- readRDS("06-11-19 beta_final.rds")
pd_final <- readRDS("06-11-19 pd_final.rds")
```


# Set the color palatte;
```{r}
library(RColorBrewer)
library(dplyr)

library(grDevices)
library(DescTools)
library(shape) #for the color legend

graphColors <- brewer.pal(8,'Set1')

graphColors1 <- brewer.pal(n =10, name = "Spectral")

graphColors2 <- c("coral",
                  "#0281e9",
                  "#64ba22",
                  "#cb4cd0",
                  "#b1d34e",
                  "#503db8",
                  "#dea800",
                  "#f59bff",
                  "#00bb6f",
                  "#a50347",
                  "#64dbae",
                  "#ff8edf",
                  "#859300",
                  "#86baff",
                  "#ff8367",
                  "#235e34",
                  "#ff94ae",
                  "#79461f",
                  "#69447e",
                  "#edbd8c",
                  "#b17291")
```


# Set up data for PCA
```{r}
#make a dataset of beta final into a pca object
prin <- prcomp(t(beta_final),center=T,scale.=F)
```


# Calculate % variance explained by each PC
```{r}
library(ggplot2)
#amount of variance explained by each PCA 1-10
screeplot(prin)
summary(prin)
#                            PC1      PC2      PC3     PC4     PC5     PC6
# Standard deviation     84.7285 20.18696 14.57732 9.16071 8.32746 8.12112
# Proportion of Variance  0.8081  0.04587  0.02392 0.00945 0.00781 0.00742
# Cumulative Proportion   0.8081  0.85396  0.87787 0.88732 0.89513 0.90255

#make a variable with Proportion of Variance for PC1-6
prop_percent <- c(0.8081, 0.04587, 0.02392, 0.00945, 0.00781, 0.00742)*100
prop_var <- data.frame(PC = c("PC1", "PC2", "PC3", "PC4", "PC5", "PC6"),
                       Percent = prop_percent)

ggplot(data = prop_var, aes(x = PC, y = Percent)) +
  geom_bar(stat = "identity")+ #identity = heights of the bars represent actual data values
        scale_y_continuous(limits=c(0, 100), breaks = seq(0, 100, 10))+
        theme(axis.title = element_text(size = 20))+
        theme_minimal()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

```


# Make the PCA - by cell type
```{r}
#output to pdf
pdf('06-17-19 PCA by cell type.pdf', width = 11, height = 8.5)
graphColors <- c("purple4", "lightsalmon1")

#make the PCA plot for cell types without whole or oragene;
plot(prin$x[ ,1],
     prin$x[ ,2],
     col = graphColors[as.factor(pd_final$celltype)],
     xlab = 'PC1', ylab = 'PC2', #axis labels
     pch = 18, #data point shape
     cex=0.5, #size of shape
     oma=c(2,2,2,13), #size of outer margin area
     xlim = c(-180, 100),
     ylim = c(-45, 35)) 
text(prin$x[,1],
     prin$x[,2],
     col = graphColors[as.factor(pd_final$celltype)],
     labels = pd_final$Sample.ID, #these are the labels
              cex=0.7, #label size
              pos = 3) 
legend('bottomleft', legend = levels(as.factor(pd_final$celltype)),
       fill = graphColors,
       title = "Cell Type")
dev.off()
```

# Make the PCA - by cell type without text
```{r}
#output to pdf
png('06-17-19 PCA by cell type - no text.png', width=1200, height=1200);
    par(mar=c(9, 14, 6, 2), mgp = c(3, 3, 0)) #mgp:(?,distance,moves axis line)
graphColors <- c("purple4", "lightsalmon1")
#make the PCA plot for cell types without whole or oragene;
plot(prin$x[ ,1],
     prin$x[ ,2],
     col = graphColors[as.factor(pd_final$celltype)],
     xlab = '', ylab = '', #axis labels
     pch = c(16, 18)[as.factor(pd_final$celltype)], #data point shape
     cex = 4, #size of shape
     cex.axis = 4,
     cex.lab = 3,
     oma=c(2,2,2,13), #size of outer margin area
     xlim = c(-180, 100),
     ylim = c(-50, 33))
mtext(text = "Principal Component 1", #this increases the distance between x-axis and text
                side = 1, #side 1 = bottom
                line = 7, cex = 4)
mtext(text = "Principal Component 2", #this increases the distance between y-axis and text
                side = 2, #side 1 = bottom
                line = 7, cex = 4)
dev.off()
```


# Make the PCA - by Sick
```{r}
#output to pdf
pdf('06-17-19 PCA by Sick.pdf', width = 11, height = 8.5)

#Color the PCA plot by Sick status
plot(prin$x[ ,1],
     prin$x[ ,2],
     col = graphColors[as.factor(pd_final$Sick)],
     xlab = 'PC1', ylab = 'PC2', #axis labels
     pch = 18, #data point shape
     cex=0.5, #size of shape
     oma=c(2,2,2,13), #size of outer margin area
     xlim = c(-180, 100),
     ylim = c(-50, 35)) 
text(prin$x[ ,1],
     prin$x[ ,2],
     col = graphColors[as.factor(pd_final$Sick)],
     labels = pd_final$Sample.ID, #these are the data labels
              cex=0.7, #label size
              pos = 3) 
legend('bottomleft', legend = levels(as.factor(pd_final$Sick)),
       fill = as.factor(pd_final$Sick) %>% as.numeric %>% as.factor %>% levels,
       title = "Sick (0 = no, 1 = yes)")
dev.off()
```


# Make the PCA - by Age Quartiles
```{r}
graphColors3 <- c("lightpink1",
                  "pink3",
                  "limegreen",
                  "slateblue3",
                  "skyblue3")

# group ages by year 7-16
#age_cat <- cut(pd_final$age.r, c(7, 8, 9, 10, 11, 12, 13, 14, 15, 16), include.lowest = TRUE)
age_quart <- round(quantile(pd_final$age.r), 2)
age_quart
#      0%      25%      50%      75%     100% 
# 7.88501 10.05065 11.70842 13.87269 16.89254 

pdf("06-17-19 PCA by Age.pdf")
plot(prin$x[ ,1], #this is PC1
     prin$x[ ,2], #this is PC2
     col = graphColors[as.factor(age_quart)],
     xlab = 'PC1', ylab = 'PC2', #axis labels
     pch = 18, #data point shape
     cex=0.5, #size of shape
     oma = c(2,2,2,13), #size of outer margin area
     xlim = c(-200, 100),
     ylim = c(-50, 40)) 
text(prin$x[,1],
     prin$x[,2],
     col = graphColors[as.factor(age_quart)],
     labels = pd_final$Sample.ID, #these are the labels
              cex=0.7, #label size
              pos = 3)
legend("bottomleft", legend = levels(as.factor(age_quart)),
       fill = graphColors[as.factor(age_quart)],
       title = "Age (years)")
dev.off()
```


# Make the PCA - by Cell count Quartiles
```{r}
#set up the quantiles
cell_quart <- round(quantile(pd_final$cell_count), 2)/10000
cell_quart
#    0%    25%    50%    75%   100% 
#  72.0  360.0  626.5 1878.0 3440.0

pdf("06-17-19 PCA by Cell Count.pdf")
plot(prin$x[ ,1], #this is PC1
     prin$x[ ,2], #this is PC2
     col = graphColors[as.factor(cell_quart)],
     xlab = 'PC1', ylab = 'PC2', #axis labels
     pch = 18, #data point shape
     cex=0.5, #size of shape
     oma = c(2,2,2,13), #size of outer margin area
     xlim = c(-200, 100),
     ylim = c(-50, 40)) 
text(prin$x[,1],
     prin$x[,2],
     col = graphColors[as.factor(cell_quart)],
     labels = pd_final$Sample.ID, #these are the labels
              cex=0.7, #label size
              pos = 3)
legend("bottomleft", legend = levels(as.factor(cell_quart)),
       fill = graphColors[as.factor(cell_quart)],
       title = "Cell Count (*10^4)")
dev.off()

```


# Subset row/slide info
```{r}
#Choose characters 1-12 within meth.id
slide <- substr(pd_final$meth_id, 1,12)
slide

#Choose the rows out of meth.id
row <- substr(pd_final$meth_id, 14,16)
row
```


# Make PC1 vs PC2 of row/slide within array
```{r}
#set colors
colors = c("yellow",
           "orange",
           "red",
           "purple",
           "blue",
           "darkgreen",
           "turquoise1",
           "sienna",
           "violetred",
           "plum2",
           "slategray")

pdf('06-29-19 PCA Row-Slide of Array.pdf')

#row
plot(prin$x[ ,1],
     prin$x[ ,2],
     col = colors, #plot by row
     xlab = 'PC1', ylab = 'PC2', #axis labels
     pch = 18, #data point shape
     cex = 0.5, #size of shape
     oma = c(2,2,2,13), #size of outer margin area
     xlim = c(-220, 100),
     ylim = c(-40, 35)) 
text(prin$x[,1],
     prin$x[,2],
     col = colors,
     labels = pd_final$Sample.ID, #these are the labels
              cex=0.7, #label size
              pos = 3) 
legend("bottomleft", legend = levels(as.factor(row)), fill = colors, title = "Row", cex = 1)

#slide
plot(prin$x[ ,1],
     prin$x[ ,2],
     col = colors,
     xlab = 'PC1', ylab = 'PC2', #axis labels
     pch = 18, #data point shape
     cex=0.5, #size of shape
     oma=c(2,2,2,13), #size of outer margin area
     xlim = c(-220, 100),
     ylim = c(-40, 35)) 
text(prin$x[,1],
     prin$x[,2],
     col = colors,
     labels = pd_final$Sample.ID, #these are the labels
              cex=0.7, #label size
              pos = 3) 
legend("bottomleft", legend = levels(as.factor(row)), fill = colors, title = "Slide", cex = 1)
dev.off()
```


# paint pc plot by quartiles of cell type
```{r}
GR_quart <- quantile(pd_final_wbc$GR, probs = c(0, 0.25, 0.5, 0.75, 1.0))
#weird rounding of quartiles
GR_quart[1] <- GR_quart[1]-0.000001

#pull the samples and they are categorical vars now
GR_q4 <- cut(pd_final_wbc$GR, breaks = GR_quart)
GR_q4

plot(prin$x[ ,1],
     prin$x[ ,2],
     col = as.factor(GR_q4),
     xlab = 'PC1', ylab = 'PC2', #axis labels
     pch = 18, #data point shape
     cex=0.5, #size of shape
     oma=c(2,2,2,13), #size of outer margin area
     xlim = c(-160, 100),
     ylim = c(-40, 35)) 
text(prin$x[,1],
     prin$x[,2],
     col = graphColors[as.factor(GR_q4)],
     labels = pd_final$Sample.ID, #these are the labels
              cex=0.7, #label size
              pos = 3) 
legend('bottomleft', legend = levels(as.factor(GR_q4)),
       fill = as.factor(GR_q4) %>% as.numeric %>% as.factor %>% levels,
       title = "Quartiles")

```



# Correlation between PC and continuous demog vars
```{r}
#PC data are in prin
#PC1 vs age
beta_final <- readRDS("06-11-19 beta_final.rds")
pd_final <- readRDS("06-11-19 pd_final.rds")
prin <- prcomp(t(beta_final),center=T,scale.=F)

#first merge in PCs into pd_final
head(prin$x) #rownames are sample which corresponds to meth_id
pc_corr <- merge(prin$x, pd_final, by.x = "row.names", by.y = "meth_id")

#Age vs PC1
cor(pc_corr$PC1, pc_corr$age.r)
#cell type vs PC1
summary(aov(pc_corr$PC1 ~ pc_corr$celltype)) #p = 1.73e-15



#set up the list
list_type <- list()

#assign 2 sublists into one list
list_type[["continuous"]] <- c("age.r",
                               "viability",
                               "cell_count",
                               "volume")
list_type[["categorical"]] <- c("celltype", "white_bin", "Male", "batch", "Sick")
list_type

#outer for loop says to run this loop 1 time
for(i in 1) 
{
  var_contin_categ <- list_type[[i]]
  print(i)
  #this loop makes pc1-6 and compares them to the demog vars from list[continuous]
  for(j in 1:6)
  {
    pc <- paste("PC", j, sep = "")
    print(pc)
    
    corr_contin <- sapply(var_contin_categ, FUN = function(var) {
  cor.test(pc_corr[,pc], pc_corr[,var])$p.value
  })
#this is stacking the dataset one row at a time
  if(j == 1) 
  {
    dataset_correlation <- corr_contin #for pc1, just make the dataset
  } else {
    dataset_correlation <- rbind(dataset_correlation, corr_contin) #after pc1, add each set of correlations as a new row
  }
  
  } 
  View(dataset_correlation)
}

#fix the rownames to be PC1-6
rownames(dataset_correlation) <- paste("PC", c(1:6), sep = "")
```


# Run categorical comparisons to PCs
```{r}
#add slide and row to pc_corr
slide <- substr(pd_final$meth_id, 1,12)
row <- substr(pd_final$meth_id, 14,16)
pc_corr$slide <- slide
pc_corr$row <- row

#sapply - categorical
#change PC# 1-6
cat_vars <- c("celltype", "white_bin", "Male", "batch", "Sick", "slide", "row")
corr_categ_pc1 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_corr$PC1 ~ pc_corr[,var]))
})
corr_categ_pc2 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_corr$PC2 ~ pc_corr[,var]))
})
corr_categ_pc3 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_corr$PC3 ~ pc_corr[,var]))
})
corr_categ_pc4 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_corr$PC4 ~ pc_corr[,var]))
})
corr_categ_pc5 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_corr$PC5 ~ pc_corr[,var]))
})
corr_categ_pc6 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_corr$PC6 ~ pc_corr[,var]))
})
```


#Set up to visualize the categorical associations
```{r}
library(gplots)
#check pc5 and batch
summary(aov(pc_corr$PC5 ~ pc_corr$batch))
boxplot(pc_corr$PC5 ~ pc_corr$batch)
stripchart(pc_corr$PC5 ~ pc_corr$batch, vertical = TRUE, method = "jitter")

#first generate matrix
p_matrix_categ <- matrix(nrow = 6, ncol = 7)
rownames(p_matrix_categ) <- c("PC1", "PC2", "PC3", "PC4", "PC5", "PC6")
colnames(p_matrix_categ) <- c("celltype",
                              "white_bin",
                              "Male",
                              "batch",
                              "Sick",
                              "Slide",
                              "Row")
View(p_matrix_categ)


#Get the info for "celltype", "white_bin", "Male", "batch", "Sick", "Slide", "Row"

#celltype, assign p-values to each cell
p_matrix_categ["PC1", "celltype"] <- corr_categ_pc1$celltype["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC2", "celltype"] <- corr_categ_pc2$celltype["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC3", "celltype"] <- corr_categ_pc3$celltype["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC4", "celltype"] <- corr_categ_pc4$celltype["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC5", "celltype"] <- corr_categ_pc5$celltype["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC6", "celltype"] <- corr_categ_pc6$celltype["pc_corr[, var]", "Pr(>F)"]

#white_bin
p_matrix_categ["PC1", "white_bin"] <- corr_categ_pc1$white_bin["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC2", "white_bin"] <- corr_categ_pc2$white_bin["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC3", "white_bin"] <- corr_categ_pc3$white_bin["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC4", "white_bin"] <- corr_categ_pc4$white_bin["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC5", "white_bin"] <- corr_categ_pc5$white_bin["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC6", "white_bin"] <- corr_categ_pc6$white_bin["pc_corr[, var]", "Pr(>F)"]
  
#Male
p_matrix_categ["PC1", "Male"] <- corr_categ_pc1$Male["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC2", "Male"] <- corr_categ_pc2$Male["pc_corr[, var]", "Pr(>F)"]  
p_matrix_categ["PC3", "Male"] <- corr_categ_pc3$Male["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC4", "Male"] <- corr_categ_pc4$Male["pc_corr[, var]", "Pr(>F)"]  
p_matrix_categ["PC5", "Male"] <- corr_categ_pc5$Male["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC6", "Male"] <- corr_categ_pc6$Male["pc_corr[, var]", "Pr(>F)"]

#batch
p_matrix_categ["PC1", "batch"] <- corr_categ_pc1$batch["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC2", "batch"] <- corr_categ_pc2$batch["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC3", "batch"] <- corr_categ_pc3$batch["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC4", "batch"] <- corr_categ_pc4$batch["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC5", "batch"] <- corr_categ_pc5$batch["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC6", "batch"] <- corr_categ_pc6$batch["pc_corr[, var]", "Pr(>F)"]

#Sick
p_matrix_categ["PC1", "Sick"] <- corr_categ_pc1$Sick["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC2", "Sick"] <- corr_categ_pc2$Sick["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC3", "Sick"] <- corr_categ_pc3$Sick["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC4", "Sick"] <- corr_categ_pc4$Sick["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC5", "Sick"] <- corr_categ_pc5$Sick["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC6", "Sick"] <- corr_categ_pc6$Sick["pc_corr[, var]", "Pr(>F)"]

#Slide
p_matrix_categ["PC1", "Slide"] <- corr_categ_pc1$slide["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC2", "Slide"] <- corr_categ_pc2$slide["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC3", "Slide"] <- corr_categ_pc3$slide["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC4", "Slide"] <- corr_categ_pc4$slide["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC5", "Slide"] <- corr_categ_pc5$slide["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC6", "Slide"] <- corr_categ_pc6$slide["pc_corr[, var]", "Pr(>F)"]

#Row
p_matrix_categ["PC1", "Row"] <- corr_categ_pc1$row["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC2", "Row"] <- corr_categ_pc2$row["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC3", "Row"] <- corr_categ_pc3$row["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC4", "Row"] <- corr_categ_pc4$row["pc_corr[, var]", "Pr(>F)"]
p_matrix_categ["PC5", "Row"] <- corr_categ_pc5$row["pc_corr[, var]", "Pr(>F)"] 
p_matrix_categ["PC6", "Row"] <- corr_categ_pc6$row["pc_corr[, var]", "Pr(>F)"]

View(p_matrix_categ)

#combine the continuous and categorical datasets of p-values
pc_demog_pvalue <- cbind(p_matrix_categ, dataset_correlation)
View(pc_demog_pvalue)
```


# Create heatmap of p-values from categ demog vars
```{r}
pdf("06-17-19 heatmap categorical demogs.pdf", height = 7, width = 11)
my_palette <- colorRampPalette(c("red", "yellow", "white"))
heatmap.2(t(pc_demog_pvalue),
          trace = "none",
          dendrogram = "none",
          key = TRUE, keysize = 1.5,
          col = my_palette,
          breaks = c(seq(0,0.1,length=100),
                         seq(0.11,0.2,length=100),
                         seq(0.21,1, length=100)),
          margins = c(5, 7),
          Rowv = FALSE,
          Colv = FALSE)
dev.off()

table(pd_final$batch, pd_final$celltype)
```


# use pairs funct to plot multiple PCs in a grid
# color has to be categorical
```{r}
library(tidyverse)
library(dplyr)
pd_final <- readRDS("06-11-19 pd_final.rds")
beta_final <- readRDS("06-11-19 beta_final.rds")
#make a dataset of beta final into a pca object
prin <- prcomp(t(beta_final),center=T,scale.=F)
age_quart <- round(quantile(pd_final$age.r), 2)
viab_quart <- round(quantile(pd_final$viability), 2)
slide <- substr(pd_final$meth_id, 1,12)
row <- substr(pd_final$meth_id, 14,16)

#set colors
colors = c("yellow",
           "orange",
           "red",
           "purple",
           "blue",
           "darkgreen",
           "turquoise1",
           "sienna",
           "violetred",
           "plum2",
           "slategray")

pdf("06-17-19 PCA demographics.pdf", height = 8.5, width = 11)
#Cell Type - FINISH
pairs(prin$x[,1:6], col = graphColors[as.factor(pd_final$celltype)], lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pd_final$celltype)), fill = graphColors, title = "Cell Type")

#Age_quart
pairs(prin$x[,1:6], col = colors[as.factor(age_quart)], lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(age_quart)), fill = colors, title = "Age (years)")

#white_bin
pairs(prin$x[,1:6], col = as.factor(pd_final$white_bin), lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pd_final$white_bin)), fill = as.factor(pd_final$white_bin) %>% as.numeric %>% as.factor %>% levels, title = "White (0=POC, 1=White)")

#Male
pairs(prin$x[,1:6], col = as.factor(pd_final$Male), lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pd_final$Male)), fill = as.factor(pd_final$Male) %>% as.numeric %>% as.factor %>% levels, title = "Male (0=Female, 1=Male)")
dim(pd_final)

#Sick
pairs(prin$x[,1:6], col = as.factor(pd_final$Sick), lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pd_final$Sick)), fill = as.factor(pd_final$Sick) %>% as.numeric %>% as.factor %>% levels, title = "Sick (0=Not Sick, 1=Sick)")

#Viability
pairs(prin$x[,1:6], col = colors[as.factor(viab_quart)], lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(viab_quart)), fill = colors, title = "Viability (%)")

#Batch
pairs(prin$x[,1:6], col = colors[as.factor(pd_final$batch)], lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pd_final$batch)), fill = colors, title = "Processing batch")

#Slide
pairs(prin$x[,1:6], col = colors[as.factor(slide)], lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(slide)), fill = colors, title = "Slide", cex = 0.8)

#Row
pairs(prin$x[,1:6], col = colors[as.factor(row)], lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(row)), fill = colors, title = "Row", cex = 1)
dev.off()
```