---
title: "06-12-19 Saliva Ttest, Ftest, Variance, Heatmaps"
author: "Lauren Middleton"
date: "6/12/2019"
output: html_document
---
```{r}
# Purpose: Run t-tests between immune and epithelial fractions, construct heatmap, calculate global methylation
# 
# Inputs:   pd_final   - details of sample fractions (epithelial and immune)
#           beta_final - beta matrix of sample fractions (epithelial and immune)
#
# Outputs:  heatmap for Figure 4b
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="C:/Users/HP/Documents/Research 2019")
knitr::opts_chunk$set(root.dir="C:/Users/HP/Documents/Research 2019")
```

# Set up;
```{r}
setwd("C:/Users/HP/Documents/Research 2019")

#load in data files from 06-11-19;
#these don't include whole or oragene or c45neg
pd_final <- readRDS("06-11-19 pd_final.rds")
beta_final <- readRDS("06-11-19 beta_final.rds")

#check that the load happened correctly;
head(pd_final)
```


# Install genefilter package from bioconductor;
```{r setup, include = FALSE}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("genefilter")
```


# Make dummy vars for cell types;
```{r}
pd_final$cd45pos <-ifelse(pd_final$celltype =="CD45pos", 1, 0)
pd_final$large <-ifelse(pd_final$celltype =="large", 1, 0)

#check the coding;
table(pd_final$celltype, pd_final$large)
table(pd_final$celltype, pd_final$cd45pos)
  #          0  1
  # CD45pos 20  0
  # large    0 18
```


# Run row ttests;
```{r}
library(genefilter)
#do a row ttest of one cell types against all other cell types within one probe;
#don't really need both but here they are
ttest_cd45pos <- rowttests(as.matrix(beta_final), as.factor(pd_final$cd45pos), tstatOnly = FALSE)
ttest_large <- rowttests(as.matrix(beta_final), as.factor(pd_final$large), tstatOnly = FALSE)
```


# How many probes are significant at 1e-8 for rowttest of one cell vs other two;
## doing lower p value because larger chance of getting <0.05
```{r}
length(which(ttest_cd45pos$p.value<1e-8)) #164793
old_probes_index <- which(x = ttest_cd45pos$p.value<1e-8)
old_probes <- rownames(ttest_cd45pos)[old_probes_index]

#for fun, what if we used 0.05 as the significance level:
length(which(ttest_large$p.value < 0.05)) #466928
466928 / 795694 #58.7%

###################################################################################################

#Adding a bonferroni cutoff for Reviewer 2 comments:

#what is the bonferroni cutoff
num_probes <- length(rownames(beta_final))
bonferroni_long <- 0.05 / num_probes
bonferroni <- round(bonferroni_long, digits = 10)

#how many probes are significant
num_sig_probes <- length(which(ttest_cd45pos$p.value < bonferroni))
#181577

#what percentage of the chip is significant
(num_sig_probes / num_probes)*100
#22.8%

#new probes
new_probes_index <- which(ttest_cd45pos$p.value < bonferroni)
new_probes <- rownames(ttest_cd45pos)[new_probes_index]

#how many of the new probes are in the old probe list
length(intersect(old_probes, new_probes))
#164793 - all of the old probes under the more strict cutoff are in the new bonferonni cutoff
```


# Subset beta_final into cell types;
```{r}
# check that order meth_ID in pd file = columns in beta matrix;
identical(colnames(beta_final), pd_final$meth_id) #TRUE

# select out the samples that are only 45pos and large cell types
cd45pos_large <- pd_final[(pd_final$celltype) %in% c("large", "CD45pos"),]
# select out the betas for these cell types - this pulls the specific columns;
cd45pos_large_beta <- beta_final[ , cd45pos_large$meth_id]
```


------------------------------------------------------------------------------------------------------
# Row ttest of one cell type vs one cell type;
```{r}
#positive vs large cells
ttest_cd45pos_large <- rowttests(as.matrix(cd45pos_large_beta),
                                 as.factor(cd45pos_large$cd45pos),
                                 tstatOnly = FALSE)
```

# How many probes are significant at 10^-8 for rowttest of one cell type vs one cell type;
#doing lower p value because larger chance of getting <0.05
#cd45- might not be a distinct cell type;
```{r}
#length(which(ttest_cd45neg_large$p.value<1e-8)) #0
length(which(ttest_cd45pos_large$p.value<1e-8)) #164793
```

------------------------------------------------------------------------------------------------------

------------------------------------------------------------------------------------------------------
# Set up for making heatmaps

# Set up for making heatmaps - Variance; 
```{r}
library(gplots)
library(genefilter) #for rowVars

#variance of every probe;
vars <- rowVars(as.matrix(beta_final))
summary(vars)
names(vars) <- rownames(beta_final)
head(vars)

#order the variance, pick out top 1000 highest variance;
vars_order <- vars[order(-vars)]
var_beta <- beta_final[rownames(beta_final) %in% names(vars_order[1:1000]),]
dim(var_beta)
```

# Histogram of variance of probes;
```{r}
hist(vars, main = "Variance of probes",
     ylab = "Count",
     xlab = "Variance",
     cex.lab = 1.5)
```


# Set up for making heatmaps - ttest; 
```{r}
#order the ttests;
#ttest_cd45neg<-ttest_cd45neg[order(ttest_cd45neg$p.value),]
ttest_cd45pos <- ttest_cd45pos[order(ttest_cd45pos$p.value), ]
ttest_large <- ttest_large[order(ttest_large$p.value), ]

#take only the top 500 ttests for each cell type;
top.t <- c(rownames(ttest_cd45pos)[1:500],
           rownames(ttest_large)[1:500])

#take out only the specified probes from top.t
top.t.beta <- beta_final[top.t, ]

#write.csv(top.t.beta, "06-12-19 top 500 ttests by cell types.csv")
```

------------------------------------------------------------------------------------------------------


------------------------------------------------------------------------------------------------------
# Make the heatmaps
```{r}
#make a heatmap;
myColors <- c("purple4",
              "lightsalmon1",
              "darkorchid",
              "firebrick1",
              "khaki1",
              "azure4",
              "black")

mydist <- function(x) dist(x,method="euclidean")

myfun <- function(x) hclust((dist((x),method="euclidean")),method="ward")

```


# Make the heatmaps of cell type specific signatures - Variance;
```{r}
#heatmap based on most variable probes
par(mar=c(2,4,4,1)+.1, oma=c(3,1,0,0))
pdf("06-12-19 Heatmap Variable Probes.pdf")
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final$celltype)],
          main="Top 500 Variable Probes",
          margins = c(8, 4))
dev.off()
```


# Make the heatmaps of t-test - edited 2/16/20
```{r}
#ttest - separates groups a little better than variance;
png("06-12-19 Top 500 Ttest.png", width=1200, height=800);
    par(mar=c(9, 9, 6, 2))
heatmap.2(as.matrix(top.t.beta), labRow=NA, labCol=NA, col=redblue(256),
          lhei = c(1,7),
          density.info="none", trace="none",
          dendrogram="column", #labCol=pd_final$Sample.ID,
          ColSideColors = myColors[as.factor(pd_final$celltype)],
          distfun = mydist, hclustfun = myfun,
          #main="Top 500 T Tests",
          key.title = "% DNA Methylation",
          key.xlab = "% DNA Methylation",
          cex.xlab = 3,
          cex.main = 3)

dev.off()
```

# Make pdf of the previous heatmaps (variance, t-test)
```{r}
#output the heatmaps as a pdf;
pdf("06-12-19 noob-Drop-Saliva-heatmap-cell-type-all.pdf")

#heatmap based on top 500 most variable probes
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final$celltype)],
          main="Top 500 Variable Probes",
          margins = c(8, 4))

#ttest - separates groups a little better than variance;
heatmap.2(as.matrix(top.t.beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none",
          dendrogram="column", labCol=pd_final$Sample.ID,
          ColSideColors = myColors[as.factor(pd_final$celltype)],
          distfun = mydist, hclustfun = myfun, main="Top 500 T Tests",
          margins = c(8,4))
dev.off()
```


# Make the heatmaps of cell type specific signatures - Variance (including person ID);
```{r}
library(gplots)
#install.packages("grDevices")

#output the heatmaps as a pdf;
pdf("06-24-19 Saliva cell types heatmap.pdf")

#heatmap based on most variable probes - age.r (blue = >12yrs)
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final$age.r>12)],
          main="Top Variable Probes - Age >12")

#heatmap based on most variable probes - Sick
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final$Sick)],
          main="Top Variable Probes - Sick")

#heatmap based on most variable probes - Race
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final$white_bin)],
          main="Top Variable Probes - Race")

#heatmap based on most variable probes - total cells
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final$cell_count>8e6)],
          main="Top Variable Probes - Total Cells")

#heatmap based on most variable probes - batch
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final$batch)],
          main="Top Variable Probes - Batch")

#heatmap based on most variable probes - viability
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final$viability>50)],
          main="Top Variable Probes - Viability >50%")

#heatmap based on most variable probes - Median Methylation Intensity - QC
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final$MQC<11.16)],
          main="Top Variable Probes - Median Methylation Intensity (QC)")

dev.off()
```

------------------------------------------------------------------------------------------------------


------------------------------------------------------------------------------------------------------

#histogram of differences in DNA meth of CD45+ vs others; - edited 2/17/20
```{r}
head(ttest_cd45pos) #dm is the difference;

#selecting only probes that meet this p value threshold;
cd45pos_sig <- ttest_cd45pos[ttest_cd45pos$p.value < 1e-08, ]

#histogram of dm from selected probes;
#multiply by -1 because the ttest is doing all vs pos instead of pos vs all;
hist(cd45pos_sig$dm*-100,
     main = "Mean Methylation Difference (p<1e-08)",
     xlim = c(-100, 100),
     xlab = "Higher methylation in Large    Higher methylation in CD45+",
     sub = "Difference in Mean Methylation between CD45+ and Large",
     cex.lab = 1.3,
     cex.sub = 1)
#more positive differences;

png("06-12-19 Mean Methylation Difference.png", width = 1600, height = 1000);
  par(mar=c(9, 11, 9, 2), mgp = c(3, 2, 0)) #mgp:(?,distance,moves axis line)
h = hist(cd45pos_sig$dm*-100, breaks=15,plot=FALSE)
ccat = cut(h$breaks, c(-Inf, -0.001, Inf))
plot(h, col=c("lightsalmon1", "purple4")[ccat],
     #main = "Mean Methylation Difference (p<1e-08)",
     main = "",
     xlim = c(-100, 100),
     xlab = "",
     ylab = "",
     #xlab = "Higher methylation in Epithelial    Higher methylation in Immune",
     #sub = "Difference in Mean Methylation between Immune and Epithelial cells",
     cex.lab = 4,
     cex.axis = 3,
     cex.sub = 4)
mtext(text = "Difference in Mean Methylation between Immune and Epithelial cells", #this increases the distance between x-axis and text
                side = 1, #side 1 = bottom
                line = 7, cex = 4)
mtext(text = "Frequency", #this increases the distance between y-axis and text
                side = 2, #side 1 = bottom
                line = 7, cex = 4)
dev.off()
```


# DNAm in epithelial vs immune
```{r}
#selecting only probes that meet p value threshold;
cd45large_sig <- ttest_large[ttest_large$p.value < 1e-08, ]
#largest magnitude of difference is 70.2% at cg07110356, p = 1.157e-22
```

# DNAm in immune vs epithelial
```{r}
#selecting only probes that meet p value threshold;
cd45pos_sig <- ttest_cd45pos[ttest_cd45pos$p.value < 1e-08, ]
#largest magnitude of difference is 68.6% at cg17804342, p = 1.611e-20
```


#histogram of differences in DNA meth of large vs others;
```{r}
head(ttest_large) #dm is the difference;

#selecting only probes that meet this p value threshold;
cd45large_sig <- ttest_large[ttest_large$p.value < 1e-08, ]

#histogram of dm from selected probes;
#multiply by -1 because the ttest is doing all vs large instead of large vs all;
hist(cd45large_sig$dm*-1)
#more negative differences;
```


# Subset saliva probes to top 1000 most different from t-test
```{r}
#load files
beta_final <- readRDS("06-11-19 beta_final.rds")
pd_final <- readRDS("06-11-19 pd_final.rds")

#create dummy vars
pd_final$cd45pos <-ifelse(pd_final$celltype =="CD45pos", 1, 0)
pd_final$large <-ifelse(pd_final$celltype =="large", 1, 0)

#do row ttest
library(genefilter)
ttest_cd45pos <- rowttests(as.matrix(beta_final), as.factor(pd_final$cd45pos), tstatOnly = FALSE)

#ttest of pos vs large
cd45pos_large <- pd_final[(pd_final$celltype) %in% c("large", "CD45pos"),]
cd45pos_large_beta <- beta_final[ , cd45pos_large$meth_id]
ttest_cd45pos_large <- rowttests(as.matrix(cd45pos_large_beta),
                                 as.factor(cd45pos_large$cd45pos),
                                 tstatOnly = FALSE)
#order the ttests;
ttest_cd45pos <- ttest_cd45pos[order(ttest_cd45pos$p.value), ]

#take only the top 500 ttests for each cell type - end up with 500;
#this is probe names
top.t <- c(rownames(ttest_cd45pos)[1:500])
#take top 500 ttest values
top.t_sep <- ttest_cd45pos[1:500, ]
table(top.t_sep$dm*-100 >0)
#468 of the top 500 are hyper meth in immune


#take out only the specified probes from top.t
top.t.beta <- beta_final[top.t, ]

#subset the saliva probes
beta_saliva_top.t <- intersect(rownames(beta_final), rownames(top.t.beta))
length(beta_saliva_top.t) #500

#keeps the order of the probes the same - grab only these probes
beta_saliva_ttest <- beta_final[beta_saliva_top.t, ]
dim(beta_saliva_ttest) #500  38

#FINISH
library(minfi)
png("07-30-19 saliva density probe subset.png", height = 500, width = 800)
densityPlot(beta_saliva_ttest,
            sampGroups = pd_final$celltype,
            pal = c("darkmagenta", "darkgreen"))
dev.off()


#only print one sample
densityPlot(beta_saliva_ttest[,5],
            sampGroups = pd_final$celltype[5,],
            pal = c("darkmagenta", "darkgreen"))
```


# Global methylation
```{r}
pd_final <- readRDS("06-11-19 pd_final.rds")
beta_final <- readRDS("06-11-19 beta_final.rds")

#put them in the same order
beta_final <- beta_final[ ,pd_final$meth_id]

#mean for each fraction
globalDNAm <- colMeans(beta_final)

#add to the pd_final
pd_final$globalDNAm <- globalDNAm

#ttest (global dnam vs cell type) for immune cells
test <- lm(globalDNAm ~ celltype, data=pd_final)
summary(test)
#                Estimate Std. Error t value Pr(>|t|)    
# (Intercept)    0.575292   0.002302  249.89  < 2e-16 ***
# celltypelarge -0.043234   0.003345  -12.93 4.34e-15 ***

#global methylation of epithelial cells
pd_new <- pd_final
pd_new$celltype <- factor(pd_final$celltype)
pd_new$celltype <- relevel(pd_new$celltype, ref = "large")

test2 <- lm(globalDNAm ~ celltype, data=pd_new)
summary(test2)
#                 Estimate Std. Error t value Pr(>|t|)    
# (Intercept)     0.532058   0.002427  219.25  < 2e-16 ***
# celltypeCD45pos 0.043234   0.003345   12.93 4.34e-15 ***
```


# Set up - Box plot of global methylation
```{r}
#run code in previous section

#make the dataset based on previous section
#57.5 s.e.: 0.2 - immune
#53.2 s.e.: 0.2 - epithelial
global_meth <- pd_final[, c("Sample.ID", "globalDNAm", "celltype")]

#make the estimated cell type values nice
global1$celltype <- gsub("CD45pos", "Immune", global_meth$celltype)
global2 <- gsub("large", "Epithelial", global1$celltype)
global_meth$celltype <- global2

#change the proportions to percents
global_meth$globalDNAm <- global_meth$globalDNAm * 100


library(ggplot2)
```


# Make Box plot of global methylation
```{r}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="C:/Users/HP/Documents/Research 2020")
knitr::opts_chunk$set(root.dir="C:/Users/HP/Documents/Research 2020")
setwd("~/Research 2020")

png("04-17-20 global DNAm boxplot.png", width=600, height=600)
ggplot(global_meth, aes(x = celltype, y = globalDNAm, fill = celltype))+ 
  geom_boxplot()+
  labs(x = "Cell Type", y = "Global DNA methylation (%)")+
  scale_y_continuous(breaks = seq(0,100,10),
                     limits = c(40,70))+ #set the y axis from 0-100 labeled every 10 units
  theme(text = element_text(size=30))+ #makes text bigger
  theme(axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))+ #change axis text color
  scale_fill_manual(values=c("lightsalmon1", "purple4"))+
  theme(panel.background = element_rect(fill = "white", #this is the background
                                colour = "black", #this is the border
                                size = 0.1, linetype = "solid"))
dev.off()
```


