---
title: "ENCODE preprocess"
author: "Lauren Middleton"
date: "7/2/2019"
output: html_document
---
```{r}
# Purpose:  QC the ENCODE data and generate a beta matrix for analysis
#
# Inputs:   meth <- "07-02-19 ENCODE ewastools-meth.rds"
#           RGset <- "07-02-19 ENCODE RGset.rds"
#           pd.encode <- "07-02-19 pd.encode.rds"
#           pd <- "07-02-19 ENCODE pd.reduced.rds" - generated in code 5.0
#
# Outputs:  "07-03-19 beta_final.rds" - beta matrix of encode/epidish cell types
#           "07-03-19 pd.paper.rds" - description of those cell types
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="C:/Users/HP/Documents/Research 2019")
knitr::opts_chunk$set(root.dir="C:/Users/HP/Documents/Research 2019")
```

```{r}
#load 06-25-19 ENCODE idats Environment.R

#if (!requireNamespace("BiocManager", quietly = TRUE))
#n    install.packages("BiocManager")

#BiocManager::install("IlluminaHumanMethylation450kmanifest")
#BiocManager::install("IlluminaHumanMethylation450kanno.ilmn12.hg19")
```

# The meth and RGSet files were saved as:
```{r}
#saveRDS(meth,file="07-02-19 ECODE ewastools-meth.rds")
#saveRDS(RGset,file="07-02-19 ENCODE RGset.rds")
#pd<-pData(RGset) #this code extracts phenotype information from an RGset, using a function from minfi
#pData(RGset)<-pd # this code applies a new phenotype table to the RGset

meth <- readRDS("07-02-19 ENCODE ewastools-meth.rds")
RGset <- readRDS("07-02-19 ENCODE RGset.rds")
pd.encode <- readRDS("07-02-19 pd.encode.rds")
pd <- readRDS("07-02-19 ENCODE pd.reduced.rds")
```


# Load the librarys
```{r}
library(ewastools)
library(minfi)
library(magrittr)
```


# Calculate QC metrics based on Illumina Guide
```{r}
#this has all of the control metrics for the array
ctrl.metrics <- control_metrics(meth)

#make a PDF of the control metric graphs
pdf("07-03-19 control_metrics.pdf")

for(metric in names(ctrl.metrics)) {
  stripchart(ctrl.metrics[[metric]],
             method = "jitter",
             pch = 4, #symbol for points
             xlab=metric)
  abline(v = attr(ctrl.metrics[[metric]],
                  "threshold"),
                  col=2, #this makes the line red
                  lty=3) } #this makes the line type dotted
dev.off()


#plot only one control metric at a time and set xlim
#https://hhhh5.github.io/ewastools/articles/exemplary_ewas.html
#staining red metric
stripchart(ctrl.metrics$"Staining Red",
           method="jitter",
           pch = 4,
           xlab = "Staining Red",
           xlim = c(0, 1500))
abline(v = 5, col = 2, lty = 3) #lty = line type

#staining green metric
stripchart(ctrl.metrics$"Staining Green",
           method="jitter",
           pch = 4,
           xlab = "Staining Green",
           xlim = c(0, 2500))
abline(v = 5, col = 2, lty = 3)

#extension green metric
stripchart(ctrl.metrics$"Extension Green",
           method = "jitter",
           pch = 4,
           xlab = "Extension Green",
           xlim = c(0, 100))
abline(v = 5, col = 2, lty = 3)
```


# Flag ctrl.metrics for any metric threshold fail
```{r}
#flag failures as TRUE if they fail on any of the 17 control metrics
fail.flag <- sample_failure(ctrl.metrics)
class(fail.flag) #logical
table(fail.flag) #2 failed, 60 passed out of 62 total samples


#FIGURE OUT WHICH SAMPLES FAILED AND ON WHICH METRICS
#WERE THESE FAILS BAD ENOUGH OR FINE

pd.flag1 <- pd.encode #this is still fine
colnames(RGset)
```


#Make a separate dataset with specific columns: geo_accession, source_name_ch1,
# characteristics_ch1.6-9
```{r}
cell_line_details <- pd.flag1[, c("geo_accession",
                                 "source_name_ch1",
                                 "characteristics_ch1.6",
                                 "characteristics_ch1.7",
                                 "characteristics_ch1.8",
                                 "characteristics_ch1.9")]

#remove "cell description: " from characteristics_ch1.6
cell_line_details$cell_description <- gsub("cell description: ",
                          replacement = "",
                          x = cell_line_details$characteristics_ch1.6)

#manually check dataset to make sure "cell description: " was removed
  #yes

#remove old characteristics_ch1.6 column
colnames(cell_line_details)
cell_line_details <- cell_line_details[-c(3)]
colnames(cell_line_details)

#export formatted dataset
#write.csv(cell_line_details, "07-02-19 ENCODE cell lines.csv")
```


# START HERE: load in "07-02-19 ENCODE preprocess Environment"
```{r}
#pd.flag has geo_accession which has all 62 samples in the format of GSM######
#sample_id in meth is under meta and starts with GSM######, but has extra words after it

#make a meth dataset that has sample_id with only the GSM###### as the ID
colnames(meth) #NULL
rownames(pd) # sample 62: "GSM999396_hg19_wgEncodeHaibMethyl450U87SitesRep1"
identical(colnames(meth), rownames(pd)) #FALSE
head(meth$meta$sample_id) #"GSM999335_hg19_wgEncodeHaibMethyl450SkmcSitesRep1"


# pd$meth_id <- gsub('_Grn.idat.gz','',pd$meth_id)
# head(pd$meth_id)
meth_gsm <- gsub('_hg19_wgEncodeHaibMethyl450.*', '', meth$meta$sample_id)
head(meth_gsm)

#add meth_gsm into meth dataset
meth$meta$gsm <- meth_gsm

#reorder the datasets so the orders match up by GSM
pd.flag <- pd.flag1[match(meth$meta$gsm, pd.flag1$geo_accession), ]

#save intial flag.fail 
pd.flag$control_probe_flag <- fail.flag
head(pd.flag)
```


# Flag for specific metric threshold fail
```{r}
#Which specific metrics fail
failed <- sapply(ctrl.metrics, function(metric) {
  metric < attr(metric, "threshold") 
})

#replace characters like spaces with underscores in failed
colnames(failed) %<>% gsub(' |/|-', '_', .)
head(failed) #now R won't break from spaces in column labels

#combine pd.flag and failed by columns
head(pd.flag)
head(failed)
pd.flag <- cbind(pd.flag, failed)
head(pd.flag)
```


# Counts of flagged samples
```{r}
#turned failed dataset into a data frame
class(failed) #matrix
failed %<>% data.frame()
class(failed) #data.frame

#add fail.flag as a new column called Any into failed dataset
failed$Any <- fail.flag

#count the column sums from failed
counts <- colSums(failed, na.rm = TRUE)
counts

#make a csv file of the control flag counts - metrics that failed
write.csv(counts, file = "07-03-19 ENCODE control.flag.counts.csv")

#there were 0 flagged samples and 2 as Any - Any means 2 samples failed
```


# Which samples were flagged
```{r}
#from pd.flag, combine geo_accession with failed
# samp.flag <- pd.flag[pd.flag$control_probe_flag,c('Sample.ID',colnames(failed)[-18])]
colnames(failed) #18 doesn't exist so I'm going to ignore that piece

#colnames of failed are the control metrics
samp.flag <- pd.flag[pd.flag$control_probe_flag, c('geo_accession', colnames(failed)[-18])]
#samples GSM999339 and GSM999368 were flagged - hepatocyte and prostate adenocarcinoma

samp.flag <- t(samp.flag)
samp.flag

write.csv(samp.flag, file = "07-03-19 ENCODE control_probe_flagged_samples.csv")

#rm(ctrl.metrics, fail.flag, failed, metric, counts)
samp.flag <- read.csv(file = "07-03-19 ENCODE control_probe_flagged_samples.csv")
```


# Correct for dye bias, check that the datasets line up
```{r}
meth %<>% correct_dye_bias()


#check pd file lines up with meth object again
rownames(pd.flag) <- pd.flag$geo_accession
identical(rownames(pd.flag),meth$meta$gsm) #TRUE
```


# Sex checks
```{r}
#X and Y normalized intensities
pd.sex <- pd.flag

#make 2 new columns for X and Y
pd.sex[ ,c('X', 'Y')] <- check_sex(meth) #this gives X and Y median intensities
head(pd.sex[ ,c('X', 'Y')])
```


```{r}
#sexes of cell lines based on pd.flag data
sex <- c("u", "u", "f", "m", "u", "f", "f", "m", "u", "m", "m", "u", "f", "f", "m", "m", "m", "m", "f", "m", "u", "u", "u", "f", "f", "u", "u", "u", "u", "u", "m", "f", "m", "m", "u", "f", "m", "m", "f", "f", "m", "f", "f", "u", "m", "u", "u", "u", "m", "u", "u", "f", "m", "f", "f", "u", "u", "f", "f", "u", "m", "f")
pd.sex$sex <- sex
colnames(pd.sex)

#predicted sex
pd.sex$predicted_sex <- predict_sex(X = pd.sex$X, Y = pd.sex$Y,
                                    male = which(pd.sex$sex == 'm'),
                                    female = which(pd.sex$sex == 'f'))
                                    #unknown = which(pd.sex$sex == "u"))

table(pd.sex$predicted_sex, pd.sex$sex)
  #    f  m  u
  # m  1 16 14
  # f 14  0  8



#which was NA or predicted sex doesn't match with data, only print specific columns
View(pd.sex[is.na(pd.sex$predicted_sex) | pd.sex$sex !=
         pd.sex$predicted_sex,c('geo_accession','sex','predicted_sex', "characteristics_ch1.7", "characteristics_ch1.8", "characteristics_ch1.9")])
#predicted sex is based on data, sex is based on annotation
#999358 = bronchial epithelial cells predict as M, written as F
```


# Plot predicted sex
```{r}
#red means predicted doesn't match with annotated including unknowns(NA)
  #if F, plot a circle etc - this is setting up the shapes
  tch <- ifelse(is.na(pd.sex$predicted_sex), 0, #if NA, plot a square
                ifelse(pd.sex$predicted_sex == "f", 1, 4)) #if F, plot circle; if M, plot X

pdf("07-02-19 sex_plot_ewastools.pdf")
  #plot X vs Y
  plot(Y ~ X, data = pd.sex,
       pch = tch, asp = 1,
      xlab = "Normalized X chromosome intensities",
      ylab = "Normalized Y chromosome intensities",
      ylim = c(0, 1.0),
      cex.lab = 1.3)
  #diff = NA or mismatch
  diff <- pd.sex[is.na(pd.sex$predicted_sex) | pd.sex$sex != pd.sex$predicted_sex,]
  #plots red mismatches or NA on top of previous points
  tch.diff <- ifelse(is.na(diff$predicted_sex), 0, 
                     ifelse(diff$predicted_sex == "f", 1, 4))
  points(Y ~ X, data = diff,
         pch = tch.diff,
         col = 2) #this makes it red ontop of plot
  legend("topright", title = "Predicted Sex", pch = c(1,4,0),legend = c("Female", "Male", "Unassigned"))
  legend("topleft", c("Sex match", "Sex mismatch"), fill = c("black", "red"))
  #text labels the points with their GSM ID
  #text(Y ~ X, labels = pd.sex$geo_accession, data = pd.sex)
dev.off()
```


#Color sex chart by cancer status of the cell line
```{r}
#first merge ENCODE cell lines dataset from csv file to pd.sex
cell_lines <- read.csv("07-02-19 ENCODE cell lines.csv")
pd.demog <- merge(pd.sex, cell_lines, by = "geo_accession")
dim(pd.demog) #62, 100


pdf("07-02-19 sex_cancer_plot_ewastools.pdf")
  #if F, plot a circle etc - this is setting up the shapes
  tch <- ifelse(is.na(pd.demog$predicted_sex), 0,
                ifelse(pd.demog$predicted_sex == "f", 1, 4))
  #plot X vs Y
  plot(Y ~ X.x, data = pd.demog,
       pch=tch,
       asp=1,
       xlab="Normalized X chromosome intensities",
       ylab="Normalized Y chromosome intensities",
       cex.lab = 1.3,
       col = ifelse(is.na(pd.demog$Karyotype), 3,
                    ifelse(pd.demog$Karyotype == "Cancer", 2, 1)))
  legend("topright", title = "Predicted Sex",
         pch = c(1, 4, 0),
         legend = c("Female", "Male", "Unassigned"))
  legend("topleft", legend = c("Cancer", "Normal", "Other Disease"),
         fill = c("red", "black", "green"))
dev.off()



saveRDS(pd.demog, file = "07-02-19 pd.demog.rds")

#saved Environment at this point
```


## Load "07-02-19 ENCODE preprocess Environment2.R"
# Detection p
```{r}
#"This function identifies failed positions defined as both the methylated and unmethylated channel reporting background signal levels."
#need to tell R to use ewastools to run this or else it gets confused because minfi and ewastools have detectionP packages that are called the exact same thing...
detP <- readRDS("07-03-19 detP.rds")

detP <- ewastools::detectionP(meth)

#make a dataset called ewastools.detP from detP on meth dataset
ewastools.detP <- detP$detP

#make a dataset called samps from the gsm names in meth
#this is the names of the samples
samps <- meth$meta$gsm
samps

#make a dataset called probes from the manifest$probe_id in detP/meth
probes <- detP$manifest$probe_id

colnames(ewastools.detP) <- samps
rownames(ewastools.detP) <- probes
print(ewastools.detP[1:6, 1:6])

# Instead, load the environment if you need detP, ewastools.detP, meth, probes, samps, pd.demog
#save(ewastools.detP, file = "07-03-19 ENCODE ewastools.detP.rda")
# load("07-03-19 ENCODE ewastools.detP.rda")

table(pd.demog$Cell.Sex, pd.demog$predicted_sex, exclude = NULL)
  #       m  f <NA>
  # F     1 14    6
  # M    19  0    3
  # <NA> 11  8    0


pdf("07-03-19 ENCODE detP_eval_ewastools.pdf")
  #this is not supposed to look like this, but could be due to the cancer cell lines
  eval_detP_cutoffs(detP, males = which(pd.demog$Cell.Sex == "M"),
                          females = which(pd.demog$Cell.Sex == "F"))
  #this is supposed to check if the mismatches change the graph
  #0.05 or 0.01 is a good cutoff for detP based on this graph
  #this plots only the lines used in the paper
  eval_detP_cutoffs(detP, females = which(pd.demog$predicted_sex == "f" & pd.demog$geo_accession %in% pd.paper$geo_accession),
                          males = which(pd.demog$predicted_sex == "m" & pd.demog$geo_accession %in% pd.paper$geo_accession))
dev.off()
```


# Assign sex based on X and Y graph
```{r}
x_assigned <- ifelse(pd.sex$Y > 0.4, "M", "F")
pd.sex$x_assigned <- x_assigned

#check if the symbols are now all assigned
  tch <- ifelse(is.na(x_assigned), 0,
                ifelse(x_assigned == "F", 1, 4))
  #plot X vs Y
  plot(Y ~ X, data = pd.sex,
       pch=tch,
       asp=1,
       xlab="Normalized X chromosome intensities",
       ylab="Normalized Y chromosome intensities",
       cex.lab = 1.3,
       col = ifelse(is.na(pd.demog$Karyotype), 3,
                    ifelse(pd.demog$Karyotype == "Cancer", 2, 1)))
  legend("topright", title = "Predicted Sex",
         pch = c(1, 4, 0),
         legend = c("Female", "Male", "Unassigned"))
  legend("topleft", legend = c("Cancer", "Normal", "Other Disease"),
         fill = c("red", "black", "green"))

```

# remake detP based on assigned sexes
```{r}
library(ewastools)
#this is based on assigned sexes
#0.05 or 0.01 is a good cutoff for detP based on this graph
  eval_detP_cutoffs(detP, males = which(pd.sex$x_assigned == "M"),
                          females = which(pd.sex$x_assigned == "F"))
  #this is based on predicted sexes
  #0.05 or 0.01 is a good cutoff for detP based on this graph
  eval_detP_cutoffs(detP, males = which(pd.demog$predicted_sex == "m"),
                          females = which(pd.demog$predicted_sex == "f"))
```



# Failed samples based on detP
```{R}
failedP <- ewastools.detP > 0.05

#Fraction of failed positions per sample
per.samp <- colMeans(failedP, na.rm=T) 
summary(per.samp)
#     Min.   1st Qu.    Median      Mean   3rd Qu.      Max. 
#0.0005272 0.0009753 0.0015015 0.0054201 0.0024253 0.1729767 

#Fraction of failed samples per position
per.probe <- rowMeans(failedP, na.rm=T) 
summary(per.probe)
#    Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
#0.000000 0.000000 0.000000 0.005421 0.000000 0.838710 

#How many samples had more than 10% of sites fail?
sum(per.samp > 0.1) # 1
length(per.samp) #62
colnames(per.samp) #NULL
#since there are no sample names in per.samp, assign names from "samps"
names(per.samp) <- samps
per.samp

#who are you
which(per.samp > 0.1)
per.samp[5] #GSM999339, 0.1729767
#this was the same primary hepatocytes


#How many positions failed >5% samples
sum(per.probe > 0.05, na.rm=T) 
# 6,612

pdf("07-03-19 ENCODE detP-hist.pdf")
  #Fraction of Failed Positions Per Sample
  hist(per.samp,breaks=20,
       xlab='Fraction of Failed Positions Per Sample',
       main='All Samples')
  #Samples with <5% Failed Probes - 
  hist(per.samp[per.samp<0.05],
       breaks=40,
       xlab='Fraction of Failed Positions Per Sample',
       main='Samples with <5% Failed Probes')
  #Fraction of Failed Samples Per Position
  hist(per.probe,breaks=20,
       xlab='Fraction of Failed Samples Per Position',
       main='All Probes')
  #Fraction of Failed Samples Per Position < 30% - zooms into left side
  hist(per.probe[per.probe<0.3],
       breaks=60,
       xlab='Fraction of Failed Samples Per Position',
       main='Probes with <30% Sampled Failed')
dev.off()

which(per.samp > 0.02) #aortic smooth muscle GSM356
```

```{r}
# save(per.probe,file='ewastools_per_probe_fail_rate.rds')
pd.detp <- pd.sex

#this is adding a new var of % probes fail per sample onto pd.detp
#this is adding the previous info from the section above into pd.detp
#rownames makes sure per.samp will be ordered the same as pd.detp
pd.detp$ew_probe_fail_pct <- per.samp[rownames(pd.detp)]
rownames(pd.detp)
# save(pd.detp,file='pd.detp.rds')

rm(ewastools.detP,failedP,per.probe,per.samp,probe.fail,probes,samps, detP)
```


# Genotype calling
```{r}
#don't normalize the beta values from meth
#simplifies meth dataset - speeds up the code
beta <- dont_normalize(meth)

#snp probes labeled as rs
snps <- meth$manifest[probe_type == 'rs', index]

#dataset of only snp probes
snps <- beta[snps, ]

#this says which bases correspond
genotypes <- call_genotypes(snps, learn=FALSE)
#saveRDS(genotypes, file = "07-03-19 called-genotypes.rds")
#genotypes <- readRDS('called-genotypes.rds')
```


# Do some checking
```{r}
#double check ordering
identical(colnames(genotypes$snps), rownames(pd.detp)) #FASLE
colnames(genotypes$snps) <- substring(colnames(genotypes$snps), 1, 9)
identical(colnames(genotypes$snps), rownames(pd.detp)) #TRUE

#check agreement of samples from same person based on snps
check.snp <- check_snp_agreement(genotypes, donor_ids = pd.detp$geo_accession,
                                 sample_ids = pd.detp$geo_accession)
# check.snp #PASTE THIS INTO THE CONSOLE



pdf("07-03-19 ENCODE SNP Agreement.pdf")
ewastools:::agreement_(genotypes, pd.detp$geo_accession, pd.detp$geo_accession)
dev.off()
```

# There are 4 pairs of cell lines that agree
## One pair is used in the paper - are these unique cell lines from 1 person or a duplicate?
```{r}
#GSM999378 and GSM999384
#SNPs fully agree with each other - do the methylation profiles agree?

#better to look at this once beta matrix has been made
#see line 801
```


# ____________________________________________________________________________________________________
# Checks with minfi - set up libraries
```{r}
library(minfi)
library(magrittr)
library(IlluminaHumanMethylation450kmanifest)
```

# Load files
```{r}
RGset <- readRDS("07-02-19 ENCODE RGset.rds")
```

# Overall intensity: M vs. U
```{r}
#pulling methylated and unmeth intensities from red and green intensities
rawMSet <- preprocessRaw(RGset)

dim(rawMSet) #dim: 485512 62, rownames are probes, colnames are samples

#M signal per probe, per sample
Meth <- getMeth(rawMSet)
Meth[1:5,1:5]

#U signal per probe, per sample
Unmeth <- getUnmeth(rawMSet)
Unmeth[1:5,1:5]

MQC <- log2(colMedians(Meth))
UQC <- log2(colMedians(Unmeth))

intensities <- data.frame(meth_id = colnames(Meth), MQC=MQC, UQC=UQC)
head(intensities) #need to shorten meth_id to match gsm
head(pd.demog) #has geo_accession

#cut off the ending of meth_id to make gsm in intensities
intensities_gsm <- gsub('_hg19_wgEncodeHaibMethyl450.*', '', intensities$meth_id)
head(intensities_gsm)
intensities$meth_id <- intensities_gsm

#merge intensities into pd.demog
pd.beta <- merge(pd.demog, intensities, by.x = "geo_accession", by.y = "meth_id")
head(pd.beta)
pd.beta$Karyotype
```


# Set up for plot of M vs U
```{r}
#replace all NAs in Karyotype with Other
Karyotype <- ifelse(is.na(pd.beta$Karyotype), "Other", as.character(pd.beta$Karyotype))
table(Karyotype)

#add Karyotype to pd.beta
pd.beta$Karyotype <- Karyotype

# Look at Cancer var
pd.beta$Karyotype <- as.factor(pd.beta$Karyotype)
class(pd.beta$Karyotype)
```


# Plot M vs U for all cell lines (n = 62)
```{r}
library(RColorBrewer)
myColors <- brewer.pal(3, "Set1")

pdf("07-03-19 M vs U plot - all cell lines.pdf")
palette(myColors)
plot(pd.beta$UQC,
     pd.beta$MQC,
     col = pd.beta$Karyotype,
     main = "M vs. U QC by Cancer Status",
     pch=16,
     xlab="Log2 Median Unmethylated Intensity",
     ylab="Log2 Median Methylated Intensity",
     cex.lab=1.2,
     cex.main=2)
legend("topleft", levels(pd.beta$Karyotype), fill = myColors)
dev.off()


pdf("07-03-19 M vs U plot - all cell lines labeled.pdf")
palette(myColors)
plot(pd.beta$UQC,
     pd.beta$MQC,
     col = pd.beta$Karyotype,
     main = "M vs. U QC by Cancer Status",
     pch=16,
     xlab="Log2 Median Unmethylated Intensity",
     ylab="Log2 Median Methylated Intensity",
     cex.lab=1.2,
     cex.main=2)
legend("topleft", levels(pd.beta$Karyotype), fill = myColors)
text(pd.beta$UQC, pd.beta$MQC, labels = pd.beta$geo_accession, pos=4, cex=0.6)
dev.off()

pdf("07-03-19 M vs U plot - all cell lines predicted_sex.pdf")
palette(myColors)
plot(pd.beta$UQC,
     pd.beta$MQC,
     col = pd.beta$predicted_sex,
     main = "M vs. U QC by Predicted Sex",
     pch=16,
     xlab="Log2 Median Unmethylated Intensity",
     ylab="Log2 Median Methylated Intensity",
     cex.lab=1.2,
     cex.main=2)
legend("topleft", levels(pd.beta$predicted_sex), fill = myColors)

dev.off()
```

# Include only the normal epithelial cell types (remove fibroblasts)
```{r}
head(pd.beta$Cell.Line)
pd.paper <- pd.beta[pd.beta$Cell.Line %in% c("HIPEpiC",
                                             "SAEC",
                                             "HRE",
                                             "HAEpiC",
                                             "HRPEpiC",
                                             "PrEC",
                                             "HEEpiC",
                                             "HCPEpiC",
                                             "HNPCEpiC",
                                             "HMEC",
                                             "HRCEpiC"
                                             # ,
                                             # "IMR90",
                                             # "ProgFib",
                                             # "AG04449",
                                             # "AG09319",
                                             # "AG04450",
                                             # "BJ",
                                             # "NHDF-neo"
                                             )
                    , ]
```


# Plot M vs U (methylated, unmethylated) of only cell lines used in paper
```{r}
library(RColorBrewer)
myColors <- brewer.pal(3, "Set1")

pdf("07-03-19 M vs U plot - paper cell lines.pdf")
palette(myColors)
plot(pd.paper$UQC,
     pd.paper$MQC,
     col = pd.paper$Karyotype,
     main = "M vs. U QC by Cancer Status",
     pch=16,
     xlab="Log2 Median Unmethylated Intensity",
     ylab="Log2 Median Methylated Intensity",
     cex.lab=1.2,
     cex.main=2)
legend("topleft", levels(pd.paper$Karyotype), fill = myColors)
dev.off()


pdf("07-03-19 M vs U plot - paper cell lines - predicted_sex.pdf")
palette(myColors)
plot(pd.paper$UQC,
     pd.paper$MQC,
     col = pd.paper$predicted_sex,
     main = "M vs. U QC by Predicted Sex",
     pch=16,
     xlab="Log2 Median Unmethylated Intensity",
     ylab="Log2 Median Methylated Intensity",
     cex.lab=1.2,
     cex.main=2)
legend("topleft", levels(pd.paper$predicted_sex), fill = myColors)
dev.off()


pdf("07-03-19 M vs U plot - paper cell lines by Cancer - labeled.pdf")
palette(myColors)
plot(pd.paper$UQC,
     pd.paper$MQC,
     col = pd.paper$Karyotype,
     main = "M vs. U QC by Cancer Status",
     pch=16,
     xlab="Log2 Median Unmethylated Intensity",
     ylab="Log2 Median Methylated Intensity",
     cex.lab=1.2,
     cex.main=2)
legend("topleft", levels(pd.paper$Karyotype), fill = myColors)
text(pd.paper$UQC, pd.paper$MQC, labels = pd.paper$geo_accession, pos=4, cex=0.6)
dev.off()
```

# Make the density plot colored by cancer
```{r}
#Raw density plot
beta.raw <- getBeta(rawMSet)

pdf("07-03-19 Density plot - Raw - all.pdf")
  densityPlot(beta.raw,
              sampGroups = pd.beta$Karyotype,
              main="Raw Beta by Normal/Cancer/Other",
              xlim = c(0.0, 1.2),
              ylim = c(0.0, 5))
dev.off()


#rm(beta.raw, intensities, Meth, MQC, Unmeth, UQC, slide, rawMSet, pd.int)
```


# Subset out only the beta.raw cell lines
```{r}
#change the labels of the columns

colnames(beta.raw)

#cut off the ending of meth_id to make gsm in intensities
raw_gsm <- gsub('_hg19_wgEncodeHaibMethyl450.*', '', colnames(beta.raw))
head(raw_gsm)
colnames(beta.raw) <- raw_gsm
identical(colnames(beta.raw), pd.beta$geo_accession) #TRUE

#subset out beta.raw to only include values for cell lines in pd.paper
beta.paper <- beta.raw[, pd.paper$geo_accession]
dim(beta.paper) #485512     11
```


# make density plot of only cell lines from paper
```{r}
pdf("07-03-19 Density plot - Raw - paper.pdf")
  densityPlot(beta.paper,
              sampGroups = pd.paper$Karyotype,
              main="Raw Beta by Normal/Cancer/Other",
              xlim = c(0.0, 1.2),
              ylim = c(0.0, 4))
dev.off()

table(pd.paper$predicted_sex)

#who is the weird wavy line
densityPlot(beta.paper,
              sampGroups = pd.paper$predicted_sex,
              main="Raw Beta by Predicted Sex",
              xlim = c(0.0, 1.2),
              ylim = c(0.0, 4))
#one of the 5 females

#subset out only the female cell lines - from pd first
pd.female <- pd.paper[pd.paper$predicted_sex == "f", ]

#subset out the beta.paper as well
beta.female <- beta.raw[, pd.female$geo_accession]
dim(beta.female)

pdf("07-03-19 Density plot - raw - female only")
densityPlot(beta.female,
              sampGroups = as.character(pd.female$Cell.Line),
              main="Raw Beta by Predicted Sex",
              xlim = c(0.0, 1.2),
              ylim = c(0.0, 4))
dev.off()
```

```{r}
rm(beta.agree, beta.female, cell_lines, pd.female, pd.agree, pd.demog, pd.sex, pd.demog)

# save these files:
saveRDS(pd.paper, file = "07-03-19 pd.paper.rds")
saveRDS(detP, file = "07-03-19 detP.rds")
saveRDS(per.probe, file = "07-03-19 per.probe.rds")
rm(pd.paper, detP, per.probe)
```



# ____________________________________________________________________________________________________
# Load "07-03-19 ENCODE preprocess Environment3.R"
```{r}
pd.paper <- readRDS("07-03-19 pd.paper.rds")
detP <- readRDS("07-03-19 detP.rds")
per.probe <- readRDS("07-03-19 per.probe.rds")
RGset <- readRDS("07-02-19 ENCODE RGset.rds")

# Load "07-02-19 ENCODE preprocess Environment3.R"
```


# Run noob correction
```{r}
library(IlluminaHumanMethylation450kmanifest)

noob <- preprocessNoob(RGset, offset=15, dyeCorr=TRUE, verbose = TRUE)

rm(RGset)

#this has all the probes and all samples
beta <- getBeta(noob)
dim(beta) #485512     62
```


# Check if the methylation profiles from the two cell lines that agreed are the same
```{r}
# this is a continuation from line 524
#compare density lines for GSM999378 and GSM999384

#subset out the cell lines that agree with each other - from pd first
#also include one that doesn't agree for comparison
pd.agree <- pd.paper[pd.paper$Cell.Line == "HNPCEpiC" |
                       pd.paper$Cell.Line == "HEEpiC" |
                       pd.paper$Cell.Line == "SAEC", ]

#mean center beta raw
mean_beta <- rowMeans(beta.raw)
mean_cent_beta <- beta.raw - mean_beta

#subset out the beta as well
beta.agree <- beta.raw[, pd.agree$geo_accession]

#compare cell lines on a density plot
densityPlot(beta.agree,
              sampGroups = as.character(pd.agree$Cell.Line),
              main="Raw Beta by Cell Line",
              xlim = c(0.0, 1.2),
              ylim = c(0.0, 4))
```

# Compare cell lines on a scatter plot
```{r}
smoothScatter(mean_cent_beta[,"GSM999378"],
              mean_cent_beta[,"GSM999384"],
              xlab = "HEEpiC",
              ylab = "HNPCEpiC")

smoothScatter(mean_cent_beta[,"GSM999359"],
              mean_cent_beta[,"GSM999392"],
              xlab = "MCF10A-1",
              ylab = "MCF10A-2")
```


# Dropping Probes

# Set up
```{r}
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)

cross_probes <- read.csv("48639-non-specific-probes-Illumina450k.csv")

data(Locations) #load locations data from illumina;
head(Locations) #X and Y chromosomes are called "chrX" or "chrY";
```

# Keep only epithelial samples
```{r}
#since the column names of beta are gsm......................, chop it down first
beta_sub <- gsub('_hg19_wgEncodeHaibMethyl450.*', '', colnames(beta))
head(beta_sub)
colnames(beta) <- beta_sub

#subset the beta because that's the corrected one
beta1 <- beta[ , pd.paper$geo_accession]
```


```{r}
failedP <- detP$detP > 0.05
failedP[1:5, 1:5] #false = not failed

#add the column names (samples) onto detP
head(detP$meta$gsm)
head(detP$manifest$probe_id)

#assign gsm to column names of failedP and probe_id to rows
colnames(failedP) <- detP$meta$gsm
rownames(failedP) <- detP$manifest$probe_id
failedP[1:5, 1:5]

#subset down to the 18 samples that were used;
failedP <- failedP[ ,pd.paper$geo_accession]
dim(failedP)

#recalculate per.probe
#probe fails in 1+ samples
per.probe <- rowMeans(failedP, na.rm = TRUE) #ignore the NAs 
table(per.probe >0.05, exclude = NULL)
# FALSE   TRUE 
#483480   2097 

#this has only the failed probes- if a probe fails in >5% it gets dropped
detp_fail <- per.probe[per.probe >0.05]

#checking the number is the same;
length(detp_fail) #2097
head(detp_fail)

#remove these two from environment because they are not used again
rm(detP, failedP)
```


```{r}
#get the probe names
detp_fail <- names(detp_fail)

#starting with this many probes;
dim(beta1) #485512     18

#keep anything that is not the failed row names;
beta_nodetp <- beta1[!rownames(beta1) %in% detp_fail, ]
dim(beta_nodetp) #480945     18


rm(beta.paper, beta.agree, beta.raw, pd.agree, RGset, rawMSet)
```


# Drop cross reactive probes;
```{r}
head(cross_probes) #Probe is an actual column here as opposed to row names;

#Number of cross reactive probes;
dim(cross_probes) #29233

#this drops all cross reactive probes;
beta_nocross <- beta_nodetp[!rownames(beta_nodetp) %in% cross_probes$"ï..TargetID", ]
dim(beta_nocross) #452023     18
```

# Dropping X and Y chromosome probes;
```{r}
#make a dataset of only X and Y;
sex_probes <- Locations[Locations$chr == "chrX" | Locations$chr == "chrY", ] 
dim(sex_probes) #11648

#check first 5 rows and columns of beta_nocross dataset;
beta_nocross[1:5, 1:5] 

#keeping rownames that is not in x and y probes;
beta_noXY <- beta_nocross[!rownames(beta_nocross) %in% rownames(sex_probes), ]

#check dimensions of dataset without x and y probes;
dim(beta_noXY) #441772
```

# Order pd.paper and beta_noXY
```{r}
#we already did the subset in the same order so this is essentially just renaming beta_noXY
beta_final <- beta_noXY[ , pd.paper$geo_accession]
dim(beta_final) #441772     18
head(beta_final)


#save the beta matrix - all probe info, rows = probes, columns = samples, data = % methylation;
saveRDS(beta_final, file = "07-03-19 beta_final.rds")

#this includes the subsetted sample data (demographic data);
saveRDS(pd.paper, file = "07-03-19 pd.paper.rds")
```


# Check PC data from ENCODE
```{r}
beta_final_encode <- readRDS("07-03-19 beta_final.rds")
pd_paper_encode <- readRDS("07-03-19 pd.paper.rds")
prin <- prcomp(t(beta_final_encode), center=T, scale.=F)
colnames(pd_paper_encode)
rownames(prin$x)
pc_pd <- merge(pd_paper_encode, prin$x, by.x = "geo_accession", by.y = "row.names")
head(pc_pd)


library(ggplot2)
#amount of variance explained by each PC
screeplot(prin)
summary(prin)
#                            PC1     PC2      PC3      PC4      PC5      PC6      PC7     PC8
# Standard deviation     39.9995 34.4552 25.36796 24.04746 21.41720 19.75899 18.58751 15.3892
# Proportion of Variance  0.2378  0.1765  0.09566  0.08596  0.06818  0.05803  0.05136  0.0352
# Cumulative Proportion   0.2378  0.4143  0.50996  0.59592  0.66410  0.72214  0.77350  0.8087

#make a variable with Proportion of Variance for PC1-8
prop_var <- data.frame(PC = c("PC1", "PC2", "PC3", "PC4", "PC5", "PC6", "PC7", "PC8"),
                       Proportion = c(0.2378,
                                      0.1765,
                                      0.09566, 0.08596, 0.06818, 0.05803, 0.05136, 0.0352))

ggplot(data = prop_var, aes(x = PC, y = Proportion)) +
  geom_bar(stat = "identity")+ #identity = heights of the bars represent actual data values
#        ylim(0, 1.0)+ #set y axis limit as 1.0
        scale_y_continuous(limits=c(0, 0.3), breaks = seq(0, 1.0, 0.1))+
        theme(axis.title = element_text(size = 20))+
        theme_minimal()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```


# Make PCA of ENCODE cell lines and WBCs
```{r}
#RUN PREVIOUS CHUNK FOR THIS TO WORK
library(magrittr)
colnames(pc_pd)
pc_pd$Cell.Type <- as.character(pc_pd$Cell.Type)
pc_pd$Cell.Lineage <- as.character(pc_pd$Cell.Lineage)

colors = c("yellow",
           "orange",
           "red",
           "purple",
           "blue",
           "darkgreen",
           "turquoise1",
           "sienna",
           "violetred",
           "plum2",
           "slategray")

#plot only the epith and fib cell lines because wbc is a separate dataset
pairs(pc_pd[, 103:110], col = as.factor(pc_pd$Cell.Type), lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pc_pd$Cell.Type)), fill = as.factor(pc_pd$Cell.Type) %>% as.numeric %>% as.factor %>% levels, title = "Cell Type")

#color by sex
pairs(pc_pd[, 103:110], col = as.factor(pc_pd$predicted_sex), lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pc_pd$predicted_sex)), fill = as.factor(pc_pd$predicted_sex) %>% as.numeric %>% as.factor %>% levels, title = "Predicted Sex")

#Color by cell lineage - THIS DOESN'T WORK - FINISH
pairs(pc_pd[, 103:110], col = as.factor(pc_pd$Cell.Lineage), lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pc_pd$Cell.Lineage)), fill = colors, title = "Cell Lineage")
table(pc_pd$Cell.Lineage, exclude = NULL)
```


# check the pc for the pairs of snp agreed cell lines
```{r}
beta_final_encode <- readRDS("07-03-19 beta_final.rds")
pd_paper_encode <- readRDS("07-03-19 pd.paper.rds")
prin <- prcomp(t(beta_final_encode), center=T, scale.=F)
colnames(pd_paper_encode)
rownames(prin$x)
pc_pd <- merge(pd_paper_encode, prin$x, by.x = "geo_accession", by.y = "row.names")
head(pc_pd)

colors = ifelse(pc_pd$geo_accession %in% c("GSM999344", "GSM999345"), "red",
                ifelse(pc_pd$geo_accession %in% c("GSM999378", "GSM999384"), "blue", "black"))

plot(pc_pd$PC1,
     pc_pd$PC2,
     col = colors,
     xlab = 'PC1', ylab = 'PC2', #axis labels
     pch = 18, #data point shape
     cex=1, #size of shape
     oma=c(2,2,2,13)) #size of outer margin area
# text(pc_pd$PC1,
#      pc_pd$PC2,
#      col = colors,
#      labels = pc_pd$geo_accession, #these are the labels
#               cex=0.7, #label size
#               pos = 3)
legend("topleft", legend = c("Fetal buttock/Fetal lung", "Esophageal/Iris", "No Agreement"), fill = c("red", "blue", "black"), title = "SNP Agreed", cex = 1)
```