---
title: "07-31-19 Reineus preprocess"
author: "Lauren Middleton"
date: "7/31/2019"
output: html_document
---
```{r}
# Purpose:  Preprocess and QC the Reinius dataset
#
# Inputs:   https://bioconductor.org/packages/release/data/experiment/html/FlowSorted.Blood.450k.html
#
# Outputs:  "07-31-19 reineus pd.qc.rds"
#           "07-31-19 reineus pd"
#           "07-31-19 reineus minfi-detP.rda"
#           "07-31-19 reineus noob.rds"
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="C:/Users/HP/Documents/Research 2019")
knitr::opts_chunk$set(root.dir="C:/Users/HP/Documents/Research 2019")
```
## Have to use minfi because starting with RGset

# Loading in the RGset of WBCs
```{r}
#loading in the RGset of WBCs
#https://bioconductor.org/packages/release/data/experiment/html/FlowSorted.Blood.450k.html
library(FlowSorted.Blood.450k)
data(FlowSorted.Blood.450k)

table(pData(FlowSorted.Blood.450k)$CellType)

#remove the unsorted samples
wbc_sorted <- FlowSorted.Blood.450k[ , !pData(FlowSorted.Blood.450k)$CellType %in% c("WBC", "PBMC", "Gran")]

table(wbc_sorted$CellType) #42 samples
# Bcell  CD4T  CD8T   Eos  Mono   Neu    NK 
#     6     6     6     6     6     6     6 
```


# Load minfi library
```{r}
library(minfi)
library(magrittr)
```


# Overall intensity: M vs. U
```{r}
#this is the equivalent of processing the RGset
rawMSet <- preprocessRaw(wbc_sorted)
rawMSet #s4 object (still includes the unsorted samples?)

#the rawMSet doesn't have the unsorted samples
table(pData(rawMSet)$CellType)

#M signal per probe, per sample
Meth <- getMeth(rawMSet)
Meth[1:5,1:5]

#U signal per probe, per sample
Unmeth <- getUnmeth(rawMSet)
Unmeth[1:5,1:5]

#find median methylation and unmethylation per sample using all probes
MQC <- log2(colMedians(Meth))
UQC <- log2(colMedians(Unmeth))

#put all the intensities into one data frame
colnames(Meth) #column names are each sample name
intensities <- data.frame(meth_id = colnames(Meth), MQC = MQC, UQC = UQC)
str(intensities) #data frame
dim(intensities) #42  3
```


# Make a pd file from the annotation info in rawMSet
```{r}
pd <- pData(rawMSet)
class(pd)
dim(pd)

head(pd)
dim(pd) #42  8


#append intensities to pd 
head(intensities) #meth_id
head(pd)
pd.int <- merge(pd, intensities, by.x = "Sample_Name", by.y = "meth_id")
dim(pd.int) #42 10
pd <- pd.int
#assign the sample names to the row names of pd
View(pd)
rownames(pd) <- pd$Sample_Name
# table(rownames(pd) %in% colnames(beta)) #42 TRUE
View(pd)

setwd("C:/Users/HP/Documents/Research 2019")
saveRDS(pd, file = "07-31-19 reineus pd.rds")
```


#plot M vs U
```{r}
#set colors
colors = c("yellow",
           "orange",
           "red",
           "purple",
           "blue",
           "darkgreen",
           "turquoise1",
           "sienna",
           "violetred",
           "plum2",
           "slategray",
           "black")

#plot M vs U colored by slide
plot(intensities$UQC, intensities$MQC,
     col = colors[as.factor(rawMSet$Slide)],
     main = "M vs. U QC by Slide",
     pch = 16,
     xlab = "Log2 Median Unmethylated Intensity", ylab="Log2 Median Methylated Intensity",
     cex.lab = 1.2, cex.main = 2)
legend("topleft", legend = levels(as.factor(rawMSet$Slide)), fill = colors, title = "Slide", cex = 0.8)

table(pd$Slide, pd$CellType)
  #            Bcell CD4T CD8T Eos Mono Neu NK
  # 5684819001     0    3    0   0    0   0  0
  # 5684819004     3    0    3   0    3   0  3
  # 5727920027     0    3    0   0    0   0  0
  # 5727920033     3    0    3   0    3   0  3
  # 5727920038     0    0    0   6    0   6  0
table(pd$Slide, pd$SampleID)
  #            043 105 149 160 218 261
  # 5684819001   0   1   0   0   1   1
  # 5684819004   0   4   0   0   4   4
  # 5727920027   1   0   1   1   0   0
  # 5727920033   4   0   4   4   0   0
  # 5727920038   2   2   2   2   2   2

#plot M vs U colored by Cell type
plot(intensities$UQC, intensities$MQC,
     col = colors[as.factor(rawMSet$CellType)],
     main = "M vs. U QC by Cell Type",
     pch = 16,
     xlab = "Log2 Median Unmethylated Intensity", ylab="Log2 Median Methylated Intensity",
     cex.lab = 1.2, cex.main = 2)
legend("topleft", legend = levels(as.factor(rawMSet$CellType)),
       fill = colors, title = "Cell Type", cex = 0.8)

#plot M vs U colored by Array
plot(intensities$UQC, intensities$MQC,
     col = colors[as.factor(rawMSet$Array)],
     main = "M vs. U QC by Array",
     pch = 16,
     xlab = "Log2 Median Unmethylated Intensity", ylab="Log2 Median Methylated Intensity",
     cex.lab = 1.2, cex.main = 2)
legend("topleft", legend = levels(as.factor(rawMSet$Array)),
       fill = colors, title = "Array", cex = 0.7)
```


# Make density plot
```{r}
#Raw density plot
beta.raw <- getBeta(rawMSet)

densityPlot(beta.raw,
            sampGroups = pd$CellType,
            main="Cell Type",
            xlim = c(0.0, 1.2),
            ylim = c(0.0, 4))

rm(beta.raw, intensities, Meth,MQC,Unmeth,UQC, rawMSet, pd.int)
```


# Calculate detP
```{r}
library(minfi)
class(wbc_sorted)
detP <- minfi::detectionP(wbc_sorted)
dim(detP) #485512     60
detP[1:5, 1:5] #samples on the columns, probes on the rows

save(detP, file = "07-31-19 reineus minfi-detP.rda")
```


# Sample and Probe fails
```{r}
load("07-31-19 reineus minfi-detP.rda")

#0.01 is arbitrary
failedP <- detP > 0.01

# Fraction of failed positions per sample
per.samp <- colMeans(failedP) 
summary(per.samp)

# Fraction of failed samples per position
per.probe <- rowMeans(failedP)
summary(per.probe)

#How many samples had more than 5% of sites fail?
sum(per.samp > 0.05) #0

# How many positions failed in at least 5% of samples? 
sum(per.probe > 0.05) #578

# How many positions failed in at least 1 samples? 
sum(per.probe > 0.01) #1057
```

# Histograms of failed probes and samples
```{r}
hist(per.samp,
     breaks=20,
     xlab='Fraction of Failed Positions Per Sample',
     main='All Samples')

hist(per.samp[per.samp<0.2],
     breaks=30,
     xlab='Fraction of Failed Positions Per Sample',
     main='Samples with <20% Failed Probes')

hist(per.probe, 
     breaks=20,
     xlab='Fraction of Failed Samples Per Position',
     main='All Probes')

hist(per.probe[per.probe<0.15],
     xlab='Fraction of Failed Samples Per Position',
     main='Probes with <15% Samples Failed')
```


# Put detp fail rate info into the pd data frame
```{r}
pd$mf_probe_fail_pct <- per.samp[pd$Sample_Name]
head(pd)

saveRDS(pd, file = "07-31-19 reineus pd.qc.rds")
```


# Noob and get betas
```{r}
noob <- preprocessNoob(wbc_sorted, offset=15, dyeCorr=TRUE, verbose = TRUE)

saveRDS(noob, file = "07-31-19 reineus noob.rds")


noob <- readRDS('07-31-19 reineus noob.rds')

beta <- getBeta(noob)

#pd <- readRDS('pd.qc.rds')
```


# 
```{r}
#assign the sample names to the row names of pd
View(pd)
rownames(pd) <- pd$Sample_Name
table(rownames(pd) %in% colnames(beta)) #42 TRUE
View(pd)

colnames(beta) #these are in a different order
pd <- pd[colnames(beta),]

identical(colnames(beta), rownames(pd)) #TRUE

```


# Dropping probes
```{r}
#skipping snp comparison because minfi can't do that

#this has only the failed probes- if a probe fails in >5% it gets dropped
detp_fail <- per.probe[per.probe >0.01]

#checking the number is the same;
length(detp_fail) #1057
head(detp_fail)

#get the probe names
detp_fail <- names(detp_fail)

#starting with this many probes;
dim(beta) #485512     42

#keep anything that is not the failed row names;
beta_nodetp <- beta[!rownames(beta) %in% detp_fail, ]
dim(beta_nodetp) #484455     42
```


# Drop cross reactive probes;
```{r}
cross_probes <- read.csv("48639-non-specific-probes-Illumina450k.csv")
head(cross_probes) #Probe is an actual column here as opposed to row names;

#Number of cross reactive probes;
dim(cross_probes) #29233

#this drops all cross reactive probes;
beta_nocross <- beta_nodetp[!rownames(beta_nodetp) %in% cross_probes$"ï..TargetID", ]
dim(beta_nocross) #455273     42
```

# Dropping X and Y chromosome probes;
```{r}
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)

data(Locations) #load locations data from illumina;
head(Locations) #X and Y chromosomes are called "chrX" or "chrY";
#make a dataset of only X and Y;
sex_probes <- Locations[Locations$chr == "chrX" | Locations$chr == "chrY", ] 
dim(sex_probes) #11648

#check first 5 rows and columns of beta_nocross dataset;
beta_nocross[1:5, 1:5] 

#keeping rownames that is not in x and y probes;
beta_noXY <- beta_nocross[!rownames(beta_nocross) %in% rownames(sex_probes), ]

#check dimensions of dataset without x and y probes;
dim(beta_noXY) #444611     42
```


# Check the order of pd and beta_noXY
```{r}
identical(rownames(pd), colnames(beta_noXY)) #TRUE
# don't need to re order anything

beta_final_wbc <- beta_noXY

#save the beta matrix - all probe info, rows = probes, columns = samples, data = % methylation;
saveRDS(beta_final_wbc, file = "07-31-19 beta_final_wbc.rds")

#this includes the subsetted sample data (demographic data);
saveRDS(pd, file = "07-31-19 reineus pd.rds")
```


# Make a PCA
```{r}
beta_final_wbc <- readRDS("07-31-19 beta_final_wbc.rds")
prin <- prcomp(t(beta_final_wbc), center = T, scale. = F)
#save(prin,file='pcs-beta.rda')

colnames(pd)
rownames(pd)
rownames(prin$x)
pc_pd <- merge(pd, prin$x, by.x = "row.names", by.y = "row.names")
head(pc_pd)

#load('pcs-beta.rda')
#pd <- readRDS('pd.qc.rds')

library(ggplot2)
#amount of variance explained by each PC
screeplot(prin)
proportion <- summary(prin)
str(proportion)
head(proportion$importance)
class(proportion$importance)
#this has the useful thing
proportion <- proportion$importance["Proportion of Variance", ]
#                            PC1     PC2      PC3      PC4     PC5 
# Standard deviation     35.1853 16.4826 14.84362 11.36295 8.45987 
# Proportion of Variance  0.5308  0.1165  0.09447  0.05536 0.03068 
# Cumulative Proportion   0.5308  0.6473  0.74173  0.79708 0.82777

#make a variable with Proportion of Variance for PC1-8
prop_var <- data.frame(PC = names(proportion[1:5]),
                       Proportion = proportion[1:5])
prop_var

ggplot(data = prop_var, aes(x = PC, y = Proportion)) +
  geom_bar(stat = "identity")+ #identity = heights of the bars represent actual data values
#        ylim(0, 1.0)+ #set y axis limit as 1.0
        scale_y_continuous(limits=c(0, 0.6), breaks = seq(0, 1.0, 0.1))+
        theme(axis.title = element_text(size = 20))+
        theme_minimal()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

```


# now do the PCA
```{r}
colors = c("yellow",
           "orange",
           "red",
           "purple",
           "blue",
           "darkgreen",
           "turquoise1",
           "sienna",
           "violetred",
           "plum2",
           "slategray")
colnames(pc_pd)

#plot wbc 
pairs(pc_pd[, 13:17], col = as.factor(pc_pd$CellType), lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pc_pd$CellType)), fill = as.factor(pc_pd$CellType) %>% as.numeric %>% as.factor %>% levels, title = "Cell Type")

#plot by slide
pairs(pc_pd[, 13:17], col = as.factor(pc_pd$Slide), lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pc_pd$Slide)), fill = as.factor(pc_pd$Slide) %>% as.numeric %>% as.factor %>% levels, title = "Slide")
```


# Run categorical comparisons to PCs
```{r}
readRDS("07-31-19 reineus pd")

colnames(pc_pd)
cat_vars <- c("Slide", "Array", "SampleID", "CellType")
cat_vars
head(pc_pd)

corr_categ_pc1 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_pd$PC1 ~ pc_pd[,var]))
})
corr_categ_pc2 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_pd$PC2 ~ pc_pd[,var]))
})
corr_categ_pc3 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_pd$PC3 ~ pc_pd[,var]))
})
corr_categ_pc4 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_pd$PC4 ~ pc_pd[,var]))
})
corr_categ_pc5 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_pd$PC5 ~ pc_pd[,var]))
})

corr_categ_pc1$Slide
```


#Set up to visualize the categorical associations
```{r}
library(gplots)
corr_categ_pc1$Slide$`Pr(>F)`

#first generate the matrix
p_matrix_categ <- matrix(nrow = 5, ncol = 4)
rownames(p_matrix_categ) <- c("PC1", "PC2", "PC3", "PC4", "PC5")
colnames(p_matrix_categ) <- c("Slide", "Array", "SampleID", "CellType")
p_matrix_categ


#Get the info for "Slide", "Array", "SampleID", "CellType"

#Slide, assign p-values to each cell
p_matrix_categ["PC1", "Slide"] <- corr_categ_pc1$Slide["pc_pd[, var]", "Pr(>F)"]
p_matrix_categ["PC2", "Slide"] <- corr_categ_pc2$Slide["pc_pd[, var]", "Pr(>F)"]
p_matrix_categ["PC3", "Slide"] <- corr_categ_pc3$Slide["pc_pd[, var]", "Pr(>F)"] 
p_matrix_categ["PC4", "Slide"] <- corr_categ_pc4$Slide["pc_pd[, var]", "Pr(>F)"] 
p_matrix_categ["PC5", "Slide"] <- corr_categ_pc5$Slide["pc_pd[, var]", "Pr(>F)"]

#Array
p_matrix_categ["PC1", "Array"] <- corr_categ_pc1$Array["pc_pd[, var]", "Pr(>F)"] 
p_matrix_categ["PC2", "Array"] <- corr_categ_pc2$Array["pc_pd[, var]", "Pr(>F)"] 
p_matrix_categ["PC3", "Array"] <- corr_categ_pc3$Array["pc_pd[, var]", "Pr(>F)"]
p_matrix_categ["PC4", "Array"] <- corr_categ_pc4$Array["pc_pd[, var]", "Pr(>F)"] 
p_matrix_categ["PC5", "Array"] <- corr_categ_pc5$Array["pc_pd[, var]", "Pr(>F)"] 
  
#SampleID
p_matrix_categ["PC1", "SampleID"] <- corr_categ_pc1$SampleID["pc_pd[, var]", "Pr(>F)"] 
p_matrix_categ["PC2", "SampleID"] <- corr_categ_pc2$SampleID["pc_pd[, var]", "Pr(>F)"]  
p_matrix_categ["PC3", "SampleID"] <- corr_categ_pc3$SampleID["pc_pd[, var]", "Pr(>F)"] 
p_matrix_categ["PC4", "SampleID"] <- corr_categ_pc4$SampleID["pc_pd[, var]", "Pr(>F)"]  
p_matrix_categ["PC5", "SampleID"] <- corr_categ_pc5$SampleID["pc_pd[, var]", "Pr(>F)"] 

#CellType
p_matrix_categ["PC1", "CellType"] <- corr_categ_pc1$CellType["pc_pd[, var]", "Pr(>F)"]
p_matrix_categ["PC2", "CellType"] <- corr_categ_pc2$CellType["pc_pd[, var]", "Pr(>F)"]
p_matrix_categ["PC3", "CellType"] <- corr_categ_pc3$CellType["pc_pd[, var]", "Pr(>F)"]
p_matrix_categ["PC4", "CellType"] <- corr_categ_pc4$CellType["pc_pd[, var]", "Pr(>F)"] 
p_matrix_categ["PC5", "CellType"] <- corr_categ_pc5$CellType["pc_pd[, var]", "Pr(>F)"]


p_matrix_categ
```


# Make heatmap of pc~demog variables
```{r}
my_palette <- colorRampPalette(c("red", "yellow", "white"))

heatmap.2(t(p_matrix_categ),
          trace = "none",
          dendrogram = "none",
          key = TRUE, keysize = 1.5,
          col = my_palette,
          breaks = c(seq(0,0.1,length=100),
                         seq(0.11,0.2,length=100),
                         seq(0.21,1, length=100)),
          margins = c(5, 7))
```
