---
title: "07-30-19 ENCODE Saliva Merge"
author: "Lauren Middleton"
date: "7/30/2019"
output: html_document
---
```{r}
# Purpose:  Merge the ENCODE, saliva, and Reinius datasets
#
# Inputs:   https://bioconductor.org/packages/release/data/experiment/html/FlowSorted.Blood.450k.html
#           "07-03-19 beta_final.rds" - encode beta matrix
#           "07-03-19 pd.paper.rds" - descriptives for encode cells
#           "06-11-19 beta_final_all_samps.rds" - beta matrix of all saliva samples
#           "06-11-19 pd_final_all_samps.rds" - descriptives for saliva samples
#           "07-31-19 reineus pd.rds" - descriptives for Reinius
#           "07-31-19 beta_final_wbc.rds" -  beta matrix of immune cells from Reinius  
#
# Outputs:  "07-31-19 reineus pd.qc.rds"
#           "07-31-19 reineus pd"
#           "07-31-19 reineus minfi-detP.rda"
#           "07-31-19 reineus noob.rds"
#           beta_overlap.rds - beta matrix that includes saliva, encode, and reinius data
#           PCA plot of all studies
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="C:/Users/HP/Documents/Research 2019")
knitr::opts_chunk$set(root.dir="C:/Users/HP/Documents/Research 2019")
```

# Drop cd45- from beta and pd
```{r}
head(pd_final_saliva)
cd45neg_samps <- pd_final_saliva$celltype == "CD45neg"
cd45neg_samps

pd_final_saliva1 <-  pd_final_saliva[!cd45neg_samps, ]
table(pd_final_saliva1$celltype)
pd_final_saliva <- pd_final_saliva1

#drop from beta
beta_final_saliva1 <- beta_final_saliva[, pd_final_saliva$meth_id]
identical(pd_final_saliva$meth_id, colnames(beta_final_saliva1))
beta_final_saliva <- beta_final_saliva1
rm(beta_final_saliva1)

saveRDS(beta_final_saliva, "06-11-19 beta_final_all_samps.rds")
```


# Load files
```{r}
#load the beta matrix - all probe info, rows = probes, columns = samples, data = % methylation;
beta_final_encode <- readRDS("07-03-19 beta_final.rds")

#this includes the subsetted sample data (demographic data);
pd_paper_encode <- readRDS("07-03-19 pd.paper.rds")

#Saliva beta final including whole and oragene
beta_final_saliva <- readRDS("06-11-19 beta_final_all_samps.rds")

#Saliva pd including whole and oragene
pd_final_saliva <- readRDS("06-11-19 pd_final_all_samps.rds")

#WBC - Reinius et al 2012
pd_final_wbc <- readRDS("07-31-19 reineus pd.rds")
beta_final_wbc <- readRDS("07-31-19 beta_final_wbc.rds")


#BiocManager::install("FlowSorted.Blood.450k")
```



# Find which probes overlap between 450k and 850k
```{r}
overlapping_probes <- Reduce(intersect, list(rownames(beta_final_encode), rownames(beta_final_saliva), rownames(beta_final_wbc)))
length(overlapping_probes) #412112 cuts out about 30,000 probes from 450k
```


# Cut betamatrix to only the ones that intersect -  subsetting to overlapping probes
```{r}
#keeps the order of the probes the same
beta_encode_overlap <- beta_final_encode[overlapping_probes, ]
beta_saliva_overlap <- beta_final_saliva[overlapping_probes, ]
beta_wbc_overlap <- beta_final_wbc[overlapping_probes, ]
dim(beta_saliva_overlap) #412112     60
dim(beta_wbc_overlap)
```


# Merge the three datasets
```{r}
beta_overlap <- cbind(beta_encode_overlap, beta_saliva_overlap, beta_wbc_overlap)
dim(beta_overlap) #410287    12
colnames(beta_overlap)

rm(beta_encode_overlap, beta_final_encode, beta_final_saliva, beta_saliva_overlap)

saveRDS(beta_overlap, file = "07-30-19 beta_overlap.rds")
```


# Combine the pds
```{r}
beta_overlap <- readRDS("07-30-19 beta_overlap.rds")


#make mini pd files with only a few overlapping variables
pd_saliva_mini <- pd_final_saliva[ , c("meth_id", "celltype", "predicted_sex")]
pd_encode_mini <- pd_paper_encode[ , c("geo_accession", "Cell.Type", "predicted_sex")]
pd_wbc_mini <- pd_final_wbc[ , c("Sample_Name", "CellType", "Sex")]


#change encode column names to match those from saliva
colnames(pd_encode_mini) <- colnames(pd_saliva_mini)
colnames(pd_wbc_mini) <- colnames(pd_saliva_mini)
head(pd_encode_mini)
head(pd_saliva_mini)
head(pd_wbc_mini)

#add study variable to each pd mini
pd_saliva_mini$study <- c("Saliva")
pd_encode_mini$study <- c("ENCODE")
pd_wbc_mini$study <- c("Reineus et al (2012)")

#combine cell types within wbc to only one category
pd_wbc_mini$celltype <- "wbc"
pd_wbc_mini$predicted_sex <- "m"

#rbind the datasets together
pd_mini <- rbind(pd_encode_mini, pd_saliva_mini, pd_wbc_mini)
pd_mini$celltype <- as.character(pd_mini$celltype)
table(pd_mini$celltype)
table(pd_mini$predicted_sex)
table(pd_mini$study)
```


# Make a PCA of samples from encode/reinius and saliva
```{r}
library(ggplot2)
#this makes the pcs
prin <- prcomp(t(beta_overlap), center=T, scale.=F)

pc_pd <- merge(pd_mini, prin$x, by.x = "meth_id", by.y = "row.names")

#how much of variance explained by pcs
screeplot(prin)
summary(prin) #PC1-7 = 80%
#                            PC1     PC2      PC3      PC4 
# Standard deviation     52.9296 30.6800 21.72696 13.45259 
# Proportion of Variance  0.4786  0.1608  0.08065  0.03092  
# Cumulative Proportion   0.4786  0.6394  0.72008  0.75100

#based on reineus file:
proportion <- summary(prin)
proportion <- proportion$importance["Proportion of Variance", ]

prop_var <- data.frame(PC = names(proportion[1:7]),
                       Proportion = proportion[1:7])

ggplot(data = prop_var, aes(x = PC, y = Proportion)) +
  geom_bar(stat = "identity")+ #identity = heights of the bars represent actual data values
#        ylim(0, 1.0)+ #set y axis limit as 1.0
        scale_y_continuous(limits=c(0, 1.0), breaks = seq(0, 1.0, 0.1))+
        theme(axis.title = element_text(size = 20))+
        theme_minimal()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```


#plot pcs of saliva+encode+wbc data
```{r}
library(dplyr)
#plot comparison of pcs 1-7
head(pc_pd)
pairs(pc_pd[, 5:11], col = as.factor(pc_pd$celltype), lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pc_pd$celltype)), fill = as.factor(pc_pd$celltype) %>% as.numeric %>% as.factor %>% levels, title = "Cell Type")

#fix the order of the legend
table(pc_pd$celltype)

#sets the order
celltype <- factor(pc_pd$celltype, levels = c("CD45pos", "large", "whole", "oragene",
                                              "Epithelial",
                                              "wbc"))


#set the colors
colors <- c("red", "deepskyblue", "purple", "deeppink1",
            "royalblue4",
            "tan")

#replot
# png("07-30-19 PCA all cell types.png", width = 700, height = 500)
# pairs(pc_pd[, 5:11], col = colors[celltype], lower.panel = NULL)
# par(xpd = TRUE)
# legend("bottomleft", legend = c("CD45+", "Large", "Whole", "Epithelial", "WBC"), fill = colors, title = "Cell Type")
# dev.off()

setwd("~/Research 2019")
#zoom in on pc1 vs 2
png("07-30-19 PC1v2 all cell types.png", width = 1000, height = 700)
plot(pc_pd$PC2,
     pc_pd$PC1,
     col = colors[celltype],
     pch = 16,
     cex = 2,
     cex.axis = 2)
par(xpd = TRUE)
# legend("topleft", legend = c("Saliva Immune", "Saliva Epithelial", "Saliva Whole", "Oragene",
#                              "ENCODE Epithelial",
#                              "Reinius Immune"), fill = colors, title = "Cell Type")
dev.off()
```

# Color by study
```{r}
png("07-30-19 PCA saliva-encode-wbc.png", width = 700, height = 500)
colors <- c("forestgreen", "gold", "deeppink2")
pairs(pc_pd[, 5:11], col = colors[as.factor(pc_pd$study)], lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pc_pd$study)),
       fill = colors,
       title = "Study")
dev.off()

pdf("07-30-19 PCA saliva-encode-wbc.pdf", width = 700, height = 500)
colors <- c("forestgreen", "gold", "deeppink2")
pairs(pc_pd[, 5:11], col = colors[as.factor(pc_pd$study)], lower.panel = NULL)
par(xpd = TRUE)
legend("bottomleft", legend = levels(as.factor(pc_pd$study)),
       fill = colors,
       title = "Study")
dev.off()
```


# epidish exploration
```{r}
library(EpiDISH)
data(centEpiFibIC.m)
colnames(centEpiFibIC.m)
rownames(centEpiFibIC.m)[1:5]
head(centEpiFibIC.m)
dim(centEpiFibIC.m) #716 3
```


# which probes are the 716 - updated 06/09/20
```{r}
# Load files and library/data
#beta matrix of QC'd data from Zheng et al - EpiDISH
beta_final_encode <- readRDS("07-03-19 beta_final.rds")
#beta matrix of QC'd data from saliva samples (includes whole/oragene)
beta_final_saliva <- readRDS("06-11-19 beta_final_all_samps.rds")
#beta matrix of QC'd data from Reinius et al - Immune cells
beta_final_wbc <- readRDS("07-31-19 beta_final_wbc.rds")
#Salvia pd including whole and oragene
pd_final_saliva <- readRDS("06-11-19 pd_final_all_samps.rds")
library(EpiDISH)
data("centEpiFibIC.m")


# Make datasets of probes that are in both the beta matrices and the epidish prediction method
dim(beta_final_encode) #There are 441772 probes in the epidish beta matrix
#make a beta matrix of probes that are in the epidish beta matrix and the ones used to predict cell proportions
beta_encode_epifibic <- intersect(rownames(beta_final_encode), rownames(centEpiFibIC.m))
length(beta_encode_epifibic) #698 of the 716 probes are in the epidish beta matrix
#make a beta matrix of probes that are in the saliva beta matrix and epidish package
beta_saliva_epifibic <- intersect(rownames(beta_final_saliva), rownames(centEpiFibIC.m))
length(beta_saliva_epifibic) #666 of the 716 probes are in the saliva beta matrix
#make a beta matrix of probes that are in the Reinius beta matrix and epidish package
beta_wbc_epifibic <- intersect(rownames(beta_final_wbc), rownames(centEpiFibIC.m)) 
length(beta_wbc_epifibic) #704 / 716

#keeps the order of the probes the same - grab only these probes
beta_encode__epifibic_overlap <- beta_final_encode[beta_encode_epifibic, ]
dim(beta_encode__epifibic_overlap) #698 18

#pull the 666 probes from the overlapping saliva and epidish datasets and make them into a matrix of rows = probes, columns = 60 saliva samples
beta_saliva_epifibic_overlap <- beta_final_saliva[beta_saliva_epifibic, ]
setwd("C:/Users/HP/Documents/Research 2019")
write.csv(beta_saliva_epifibic_overlap, "saliva_epidish_overlap.csv")

#pull the same "716" probes from wbc
beta_wbc_epifibic_overlap <- beta_final_wbc[beta_wbc_epifibic, ]

#plot density plot of each
library(minfi)
identical(pd_final_saliva$meth_id, colnames(beta_saliva_epifibic_overlap))
png("07-30-19 saliva density probe encode subset.png")
densityPlot(beta_saliva_epifibic_overlap, sampGroups = pd_final_saliva$celltype)
dev.off() #saliva
densityPlot(beta_encode__epifibic_overlap, sampGroups = pd_final_encode$celltype) #encode
```


# Test for associations between pcs and variables - these numbers are old-(8/5/20)
```{r}
#RUN COMBINE THE PDS SECTION FIRST

#this makes the pcs
prin <- prcomp(t(beta_overlap), center=T, scale.=F)
pc_pd <- merge(pd_mini, prin$x, by.x = "meth_id", by.y = "row.names")
#pc_pd has the PCs and celltype/study columns

#Study and celltype vs PC1 and PC2
cat_vars <- c("celltype", "study")
corr_categ_pc1 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_pd$PC1 ~ pc_pd[,var]))
})
corr_categ_pc2 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_pd$PC2 ~ pc_pd[,var]))
})
corr_categ_pc3 <- sapply(cat_vars, FUN = function(var) {
  summary(aov(pc_pd$PC3 ~ pc_pd[,var]))
})

#PC1:
corr_categ_pc1
corr_categ_pc1[[1]]$'Pr(>F)'
#cell type is associated with PC1 (p=1.8e-50)
#study is also associated with PC1 (p=4.3e-27)

#PC2:
corr_categ_pc2
#cell type is associated with PC2 (p=2.3e-51)
#same for study (p=6.1e-48)

#PC3:
#also associated, celltype associated with PC4
```