---
title: "11-21-19 Saliva estimation"
author: "Lauren Middleton"
date: "11/21/2019"
output: html_document
---
```{r}
# Purpose:  graph the ewastools estimates of saliva samples - some figures and stats were updated in code 7.1
#
# Inputs:  "01-23-20 estimates_subset_id.csv" 
#
# Outputs:  plots and stats
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="C:/Users/HP/Documents/Research 2020")
knitr::opts_chunk$set(root.dir="C:/Users/HP/Documents/Research 2020")
```


# Load in data and estimate with ewastools and saliva ref panel, fix things
```{r}
#run ewastools package with saliva reference panel
library(ewastools)

# Load datasets
setwd("C:/Users/HP/Documents/Research 2019")
pd_final_all <- readRDS("06-11-19 pd_final_all_samps.rds")
beta_final_all <- readRDS("06-11-19 beta_final_all_samps.rds")

out.1 <- estimateLC(meth = beta_final_all, ref = "lauren_saliva_ref")
head(out.1)

#make it a dataframe
est <- as.data.frame(out.1)
# View(est)

#when package is live, use estimateLC from ewastools to get the cell proportion estimates:
# estimates_subset_id <- estimateLC(beta_final, "saliva")

#add sample names and cell types
est$sample <- pd_final_all$Sample.ID
est$celltype <- pd_final_all$celltype
table(est$celltype)
# CD45pos   large oragene   whole 
#      20      18       4      18

#make the cell type names nice
est$celltype <- gsub("large", "Epithelial Fractions", est$celltype)
est$celltype <- gsub("CD45pos", "Immune Fractions", est$celltype)
est$celltype <- gsub("whole", "Whole Samples", est$celltype)
est$celltype <- gsub("oragene", "Oragene Kits", est$celltype)
table(est$celltype)

#make the dataset into the long form
library(tidyverse)
saliva_long <- gather(est, cell_type_est, estimate, IC:Epi, factor_key=TRUE)

#change proportions into percentages
saliva_long$estimate <- saliva_long$estimate*100

table(saliva_long$cell_type_est) #60 of each
saliva_long$cell_type_est <- gsub("Epi", "Epithelial", saliva_long$cell_type_est)
saliva_long$cell_type_est <- gsub("IC", "Immune", saliva_long$cell_type_est)
table(saliva_long$cell_type_est) #60 of each

# View(saliva_long)
```


# Subset full dataset into cell types
```{r}
table(saliva_long$celltype)
# Epithelial Fractions     Immune Fractions         Oragene Kits        Whole Samples 
#                   36                   40                    8                   36 

#subset saliva_long into Immune
immune_est <- saliva_long[saliva_long$celltype == "Immune Fractions", ]
head(immune_est)
table(immune_est$celltype)
# Immune Fractions 
#               40 

#subset saliva_long into Epithelial
epithelial_est <- saliva_long[saliva_long$celltype == "Epithelial Fractions", ]
head(epithelial_est)
table(epithelial_est$celltype)

#subset saliva_long into Oragene
oragene_est <- saliva_long[saliva_long$celltype == "Oragene Fractions", ]
head(oragene_est)
table(oragene_est$celltype)

#subset saliva_long into Whole
whole_est <- saliva_long[saliva_long$celltype == "Whole Fractions", ]
head(whole_est)
table(whole_est$celltype)
```


# Graph Immune estimates as spaghetti plot
```{r}
pdf("01-31-20 immune fraction ewastools spaghetti prediction.pdf")
ggplot(data = immune_est, aes(x = cell_type_est, y = estimate, group = sample))+
  geom_point(aes(color = sample), size = 3)+
  geom_line(aes(color = sample), size = 1)+
  xlab("")+
  ylab("Percent Cell Type (%)")+
  scale_y_continuous(breaks = seq(0,120,10),
                     limits = c(0,120))+ #set the y axis from 0-100 labeled every 10 units
  theme(text = element_text(size=20))+ #makes text bigger
  theme(axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))+ #change axis text color
  theme(panel.background = element_rect(fill = "white", #this is the background
                                colour = "black", #this is the border
                                size = 0.1, linetype = "solid"))+
  theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
dev.off()
```


# Graph Immune predictions as violin plot
```{r}
pdf("01-31-20 immune fraction ewastools prediction.pdf")
ggplot(data = immune_est, aes(x = cell_type_est, y = estimate, fill = cell_type_est))+
  geom_violin()+
  geom_point(size = 2)+
  xlab("")+
  ylab("Percent Cell Type (%)")+
  labs(title = "Immune fractions")+
  scale_y_continuous(breaks = seq(0,120,10),
                     limits = c(0,120))+ #set the y axis from 0-100 labeled every 10 units
  theme(text = element_text(size=20))+ #makes text bigger
  theme(axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))+ #change axis text color
  theme(panel.background = element_rect(fill = "white", #this is the background
                                colour = "black", #this is the border
                                size = 0.1, linetype = "solid"))+
  theme(legend.position="bottom")+
  theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
dev.off()
```


# Graph Epithelial predictions as violin plot
```{r}
pdf("01-31-20 epithelial fraction ewastools prediction.pdf")
ggplot(data = epithelial_est, aes(x = cell_type_est, y = estimate, fill = cell_type_est))+
  geom_violin()+
  geom_point(size = 2)+
  xlab("")+
  ylab("Percent Cell Type (%)")+
  labs(title = "Epithelial fractions")+
  scale_y_continuous(breaks = seq(0,120,10),
                     limits = c(0,120))+ #set the y axis from 0-100 labeled every 10 units
  theme(text = element_text(size=20))+ #makes text bigger
  theme(axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))+ #change axis text color
  theme(panel.background = element_rect(fill = "white", #this is the background
                                colour = "black", #this is the border
                                size = 0.1, linetype = "solid"))+
  theme(legend.position="bottom")+
  theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
dev.off()
```


# Graph Oragene predictions as violin plot
```{r}
pdf("01-31-20 oragene fraction ewastools prediction.pdf")
ggplot(data = oragene_est, aes(x = cell_type_est, y = estimate, fill = cell_type_est))+
  geom_violin()+
  geom_point(size = 2)+
  xlab("")+
  ylab("Percent Cell Type (%)")+
  labs(title = "Oragene samples")+
  scale_y_continuous(breaks = seq(0,120,10),
                     limits = c(0,120))+ #set the y axis from 0-100 labeled every 10 units
  theme(text = element_text(size=20))+ #makes text bigger
  theme(axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))+ #change axis text color
  theme(panel.background = element_rect(fill = "white", #this is the background
                                colour = "black", #this is the border
                                size = 0.1, linetype = "solid"))+
  theme(legend.position="bottom")+
  theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
dev.off()
```


# Graph Whole predictions as violin plot
```{r}
pdf("01-31-20 whole fraction ewastools prediction.pdf")
ggplot(data = whole_est, aes(x = cell_type_est, y = estimate, fill = cell_type_est))+
  geom_violin()+
  geom_point(size = 2)+
  xlab("")+
  ylab("Percent Cell Type (%)")+
  labs(title = "Whole samples")+
  scale_y_continuous(breaks = seq(0,120,10),
                     limits = c(0,120))+ #set the y axis from 0-100 labeled every 10 units
  theme(text = element_text(size=20))+ #makes text bigger
  theme(axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))+ #change axis text color
  theme(panel.background = element_rect(fill = "white", #this is the background
                                colour = "black", #this is the border
                                size = 0.1, linetype = "solid"))+
  theme(legend.position="bottom")+
  theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
dev.off()
```



# Make a panel figure of the four graphs - remade in code 7.1
```{r}
saliva_long$cell_type_est <- factor(saliva_long$cell_type_est,
                                      levels = c("Immune", "Epithelial"))

png("07-29-20 saliva estimates panel.png", width=1000, height=1000)
ggplot(data = saliva_long, aes(x = cell_type_est, y = estimate, fill = cell_type_est))+
  geom_violin(width = 0.5)+
  #geom_point(size = 2)+
  geom_jitter(width = 0.05, size = 3)+
  xlab("")+
  ylab("Percent Cell Type (%)")+
  scale_y_continuous(breaks = seq(0,120,10),
                     limits = c(-1,120))+ #set the y axis from 0-100 labeled every 10 units
  theme(text = element_text(size=35))+ #makes text bigger
  labs(fill = "Estimated Cell Types")+ #change the legend title
  theme(axis.text.x = element_text(colour = "black"),
        axis.text.y = element_text(colour = "black"))+ #change axis text color
  theme(panel.background = element_rect(fill = "white", #this is the background
                                colour = "black", #this is the border
                                size = 0.1, linetype = "solid"))+
  theme(legend.position="bottom")+
  theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())+
  facet_wrap(vars(celltype))
dev.off()
```

# Stats - updated in code 7.1