---
title: "06-19-19 WBC type estimation"
author: "Lauren Middleton"
date: "6/19/2019"
output: html_document
---
```{r}
# Purpose:  estimate the proportions of white blood cells in the saliva samples using Reinius reference panel in ewastools
#
# Inputs:   "06-11-19 pd_final_all_samps.rds" - descriptives of saliva samples
#           "06-11-19 beta_final_all_samps.rds" - beta matrix from saliva samples
#
# Outputs:  plots and basic stats
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="C:/Users/HP/Documents/Research 2019")
knitr::opts_chunk$set(root.dir="C:/Users/HP/Documents/Research 2019")
```

# Load everything
```{r}
#devtools::install_github("hhhh5/ewastools")

library(ewastools)
library(genefilter) #for rowVars
setwd("C:/Users/HP/Documents/Research 2019")

#load in data files from 06-11-19;
pd_final_all <- readRDS("06-11-19 pd_final_all_samps.rds")
beta_final_all <- readRDS("06-11-19 beta_final_all_samps.rds")

#instead load "06-19-19 07-08-19 Saliva WBC environment.RData"

#check that the load happened correctly;
head(pd_final_all)
```

# Do the proportion estimation and do some logic checks
```{r}
cells_comp <- estimateLC(beta_final_all, "Reinius")

class(cells_comp) #data table

#turning cells_comp from a data table into a data frame
cells_comp <- data.frame(cells_comp)

#attaching our sample IDs to the data frame instead of the numbers it was
rownames(cells_comp) <- colnames(beta_final_all)
head(cells_comp)

#how many predicted as negative
table(cells_comp$CD8 < 0) #5 predict as 0
table(cells_comp < 0) #1 predicts as 0 outside of CD8
table(cells_comp == 0)

#reassign the negative values as 0
cells_comp[cells_comp <0] <- 0

#how many predict as >1 (this is a proportion, not %)
table(cells_comp$GR > 1) #2 predict as >100

#reassign the 1+ values as 1
cells_comp[cells_comp >1] <- 1
```


# Merge cell composition data frame with pd_final_all
```{r}
# match meth ID to meth ID in pd final to merge the two
wbc <- merge(pd_final_all, cells_comp, by.x = "meth_id", by.y = "row.names")
dim(wbc)
pd_final_all_wbc <- wbc
```

## Main processing is done ##
#____________________________________________________________________________________________________#

# Calculate ranges of WBCs in saliva samples
```{r}
table(pd_final_all_wbc$GR, pd_final_all_wbc$celltype == "CD45pos")
pd_final_all_wbc$GR

#calculate the ranges of predicted wbcs

#Granulocytes:
tapply(pd_final_all_wbc$GR, pd_final_all_wbc$celltype, range)
#immune: 30.5-100%
#epith: 12.2-24.6%
#oragene: 41.1-86.4%
#whole: 19.5-100%

#CD4:
tapply(pd_final_all_wbc$CD4, pd_final_all_wbc$celltype, range)
#immune: 0-28.0%
#epith: 35.0-42.3%
#oragene: 6.8-28.1%
#whole: 0.5-39.4%

#CD8:
tapply(pd_final_all_wbc$CD8, pd_final_all_wbc$celltype, range)
#immune: 0%
#epith: 0%
#oragene: 0%
#whole: 0%

#B cells
tapply(pd_final_all_wbc$B, pd_final_all_wbc$celltype, range)
#immune: 3.3-20.0%
#epith: 18.2-20.7%
#oragene: 8.8-13.8%
#whole: 0.5-39.4%

#Monocytes
tapply(pd_final_all_wbc$MO, pd_final_all_wbc$celltype, range)
#immune: 0-14.1%
#epith: 16.5-20.6%
#oragene: 0-11.3%
#whole: 0.2-16.8%

#Natural Killer
tapply(pd_final_all_wbc$NK, pd_final_all_wbc$celltype, range)
#immune: 0-8.7%
#epith: 5.8-8.8%
#oragene: 1.9-6.1%
#whole: 0-8.3%
```

## 3/3/20
# Make violin plots with all WBC types on one plot for each saliva fraction

# Planning
```{r}
View(pd_final_all_wbc)
#celltype has the fraction types - might have to select out the CD45pos ones first
#row is each sample
#there are columns for each WBC type - these are in decimals - convert to %
table(pd_final_all_wbc$celltype)

#PLAN
#step 1: keep only the columns with Sample_Group(column 2), celltype(column 11), and WBCs(columns 65-70)
#step 2: make a new dataset with only immune cell types - should be 20 rows x 8 columns
#step 3: turn the WBC columns wide to long - check epidish file for code
#step 4: turn decimal estimates into percents
#step 5: fix names of GR:NK to Granulocyte:Natural Killer
#step 6: graph using ggplot2
```

# Set up the immune dataset
```{r}
#step 1: keep only the columns with Sample_Group(column 2), celltype(column 11), and WBCs(columns 85-90)
wbc_all_samp <- pd_final_all_wbc[ , c(2, 11, 65:70)]
colnames(wbc_all_samp) #60 obs, 8 vars

#step 2: make a new dataset with only immune cell types - should be 20 rows x 8 columns
immune_wbc <- wbc_all_samp[wbc_all_samp$celltype == "CD45pos", ]
table(immune_wbc$celltype) #20 CD45pos observations

#step 3: turn the WBC columns wide to long - check epidish file for code
library(tidyr)
#gather = function
#immune_wbc = original dataset
#wbc_cell_type = column with wbc cell types
#estimate = proportion estimate numbers
#GR:NK = columns to turn from wide to long
immune_wbc_est_long <- gather(immune_wbc, wbc_cell_type, estimate, GR:NK, factor_key=TRUE)
View(immune_wbc_est_long)

#step 4: turn decimal estimates into percents
immune_wbc_est_long$estimate <- immune_wbc_est_long$estimate*100

#step 5: fix names of GR:NK to Granulocyte:Natural Killer
table(immune_wbc_est_long$wbc_cell_type) #20 of each wbc type
 # GR CD4 CD8   B  MO  NK 
 # 20  20  20  20  20  20
immune_wbc_est_long$wbc_cell_type <- gsub("GR", "Granulocyte", immune_wbc_est_long$wbc_cell_type)
immune_wbc_est_long$wbc_cell_type <- gsub("B", "B cell", immune_wbc_est_long$wbc_cell_type)
immune_wbc_est_long$wbc_cell_type <- gsub("MO", "Monocyte", immune_wbc_est_long$wbc_cell_type)
immune_wbc_est_long$wbc_cell_type <- gsub("NK", "Natural Killer", immune_wbc_est_long$wbc_cell_type)
table(immune_wbc_est_long$wbc_cell_type)

range(immune_wbc$GR) #0.3-1.0
```

# Set up the Whole saliva dataset
```{r}
#step 1: keep only the columns with Sample_Group(column 2), celltype(column 11), and WBCs(columns 85-90)
wbc_all_samp <- pd_final_all_wbc[ , c(2, 11, 65:70)]
colnames(wbc_all_samp) #60 obs, 8 vars

#step 2: make a new dataset with only immune cell types - should be 20 rows x 8 columns
whole_wbc <- wbc_all_samp[wbc_all_samp$celltype == "whole", ]
table(whole_wbc$celltype) #18 Whole observations

#step 3: turn the WBC columns wide to long - check epidish file for code
library(tidyr)
#gather = function
#immune_wbc = original dataset
#wbc_cell_type = column with wbc cell types
#estimate = proportion estimate numbers
#GR:NK = columns to turn from wide to long
whole_wbc_est_long <- gather(whole_wbc, wbc_cell_type, estimate, GR:NK, factor_key=TRUE)
View(whole_wbc_est_long)

#step 4: turn decimal estimates into percents
whole_wbc_est_long$estimate <- whole_wbc_est_long$estimate*100

#step 5: fix names of GR:NK to Granulocyte:Natural Killer
table(whole_wbc_est_long$wbc_cell_type) #18 of each wbc type

whole_wbc_est_long$wbc_cell_type <- gsub("GR", "Granulocyte", whole_wbc_est_long$wbc_cell_type)
whole_wbc_est_long$wbc_cell_type <- gsub("B", "B cell", whole_wbc_est_long$wbc_cell_type)
whole_wbc_est_long$wbc_cell_type <- gsub("MO", "Monocyte", whole_wbc_est_long$wbc_cell_type)
whole_wbc_est_long$wbc_cell_type <- gsub("NK", "Natural Killer", whole_wbc_est_long$wbc_cell_type)
table(whole_wbc_est_long$wbc_cell_type)
```


# Set up the Oragene dataset
```{r}
#step 1: keep only the columns with Sample_Group(column 2), celltype(column 11), and WBCs(columns 85-90)
wbc_all_samp <- pd_final_all_wbc[ , c(2, 11, 65:70)]
colnames(wbc_all_samp) #60 obs, 8 vars

#step 2: make a new dataset with only immune cell types - should be 20 rows x 8 columns
oragene_wbc <- wbc_all_samp[wbc_all_samp$celltype == "oragene", ]
table(oragene_wbc$celltype) #4 oragene observations

#step 3: turn the WBC columns wide to long - check epidish file for code
library(tidyr)
#gather = function
#immune_wbc = original dataset
#wbc_cell_type = column with wbc cell types
#estimate = proportion estimate numbers
#GR:NK = columns to turn from wide to long
oragene_wbc_est_long <- gather(oragene_wbc, wbc_cell_type, estimate, GR:NK, factor_key=TRUE)
View(oragene_wbc_est_long)

#step 4: turn decimal estimates into percents
oragene_wbc_est_long$estimate <- oragene_wbc_est_long$estimate*100

#step 5: fix names of GR:NK to Granulocyte:Natural Killer
table(oragene_wbc_est_long$wbc_cell_type) #4 of each wbc type

oragene_wbc_est_long$wbc_cell_type <- gsub("GR", "Granulocyte", oragene_wbc_est_long$wbc_cell_type)
oragene_wbc_est_long$wbc_cell_type <- gsub("B", "B cell", oragene_wbc_est_long$wbc_cell_type)
oragene_wbc_est_long$wbc_cell_type <- gsub("MO", "Monocyte", oragene_wbc_est_long$wbc_cell_type)
oragene_wbc_est_long$wbc_cell_type <- gsub("NK", "Natural Killer", oragene_wbc_est_long$wbc_cell_type)
table(oragene_wbc_est_long$wbc_cell_type)
```


# Graphing immune fraction WBC estimates
```{r}
#step 6: graph using ggplot2
library(ggplot2)


immune_wbc_est_long$wbc_cell_type <- factor(immune_wbc_est_long$wbc_cell_type, levels = c("Granulocyte", "CD4", "CD8", "B cell", "Monocyte", "Natural Killer"))

png("06-19-19 03-03-20 WBC prediction from Reinius - Immune.png", width=1300, height=1000)
ggplot(data = immune_wbc_est_long, aes(x = wbc_cell_type, y = estimate, fill = wbc_cell_type))+
  geom_violin(scale="count", width=1.5)+
  #GR
  geom_rect(aes(xmin = 0.5, xmax = 1.5,
                ymin = 38, ymax = 72),
                fill = "pink", alpha = 0.02)+
  #CD4
  geom_rect(aes(xmin = 1.5, xmax = 2.5,
                ymin = 31, ymax = 52),
                fill = "pink", alpha = 0.02)+
  #CD8
  geom_rect(aes(xmin = 2.5, xmax = 3.5,
                ymin = 18, ymax = 35),
                fill = "pink", alpha = 0.02)+
  #B
  geom_rect(aes(xmin = 3.5, xmax = 4.5,
                ymin = 11.3, ymax = 18),
                fill = "pink", alpha = 0.02)+ 
  #MO
  geom_rect(aes(xmin = 4.5, xmax = 5.5,
                ymin = 0, ymax = 8),
                fill = "pink", alpha = 0.02)+  
  #NK
  geom_rect(aes(xmin = 5.5, xmax = 6.5,
                ymin = 4, ymax = 27),
                fill = "pink", alpha = 0.02)+  
  geom_jitter(width = 0.08, size = 3, aes(size = 3))+
  xlab("")+
  ylab("Percent Cell Type (%)")+
  scale_y_continuous(breaks = seq(0,100,10),
                     limits = c(-1,100))+ #set the y axis from 0-100 labeled every 10 units
  theme(text = element_text(size=35))+ #makes text bigger
  labs(fill = "Estimated Cell Types")+ #change the legend title
  theme(axis.text.x = element_text(colour = "black", size = 35),
        axis.text.y = element_text(colour = "black"))+ #change axis text color
  theme(panel.background = element_rect(fill = "white", #this is the background
                                colour = "black", #this is the border
                                size = 0.1, linetype = "solid"))+
  #scale_fill_manual(values=c("#F8766D", "#00BFC4", "yellow"))+
  theme(legend.position="none")+
  theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
dev.off()
```

# Graphing Whole fraction WBC estimates
```{r}
#step 6: graph using ggplot2
library(ggplot2)


whole_wbc_est_long$wbc_cell_type <- factor(whole_wbc_est_long$wbc_cell_type, levels = c("Granulocyte", "CD4", "CD8", "B cell", "Monocyte", "Natural Killer"))

png("06-19-19 03-03-20 WBC prediction from Reinius - Whole.png", width=1300, height=1000)
ggplot(data = whole_wbc_est_long, aes(x = wbc_cell_type, y = estimate, fill = wbc_cell_type))+
  geom_violin(scale="count", width=1.5)+
  #GR
  geom_rect(aes(xmin = 0.5, xmax = 1.5,
                ymin = 38, ymax = 72),
                fill = "pink", alpha = 0.02)+
  #CD4
  geom_rect(aes(xmin = 1.5, xmax = 2.5,
                ymin = 31, ymax = 52),
                fill = "pink", alpha = 0.02)+
  #CD8
  geom_rect(aes(xmin = 2.5, xmax = 3.5,
                ymin = 18, ymax = 35),
                fill = "pink", alpha = 0.02)+
  #B
  geom_rect(aes(xmin = 3.5, xmax = 4.5,
                ymin = 11.3, ymax = 18),
                fill = "pink", alpha = 0.02)+ 
  #MO
  geom_rect(aes(xmin = 4.5, xmax = 5.5,
                ymin = 0, ymax = 8),
                fill = "pink", alpha = 0.02)+  
  #NK
  geom_rect(aes(xmin = 5.5, xmax = 6.5,
                ymin = 4, ymax = 27),
                fill = "pink", alpha = 0.02)+   
  geom_jitter(width = 0.08, size = 3, aes(size = 3))+
  xlab("")+
  ylab("Percent Cell Type (%)")+
  scale_y_continuous(breaks = seq(0,100,10),
                     limits = c(-1,100))+ #set the y axis from 0-100 labeled every 10 units
  theme(text = element_text(size=35))+ #makes text bigger
  labs(fill = "Estimated Cell Types")+ #change the legend title
  theme(axis.text.x = element_text(colour = "black", size = 35),
        axis.text.y = element_text(colour = "black"))+ #change axis text color
  theme(panel.background = element_rect(fill = "white", #this is the background
                                colour = "black", #this is the border
                                size = 0.1, linetype = "solid"))+
  #scale_fill_manual(values=c("#F8766D", "#00BFC4", "yellow"))+
  theme(legend.position="none")+
  theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())

dev.off()
```


# Graphing Oragene fraction WBC estimates
```{r}
#step 6: graph using ggplot2
library(ggplot2)


oragene_wbc_est_long$wbc_cell_type <- factor(oragene_wbc_est_long$wbc_cell_type,
                                             levels = c("Granulocyte",
                                                        "CD4",
                                                        "CD8",
                                                        "B cell",
                                                        "Monocyte",
                                                        "Natural Killer"))

png("06-19-19 03-03-20 WBC prediction from Reinius - Oragene.png", width=1300, height=1000)
ggplot(data = oragene_wbc_est_long, aes(x = wbc_cell_type, y = estimate, fill = wbc_cell_type))+
  geom_violin(scale="count", width=1.5)+
  #GR
  geom_rect(aes(xmin = 0.5, xmax = 1.5,
                ymin = 38, ymax = 72),
                fill = "pink", alpha = 0.02)+
  #CD4
  geom_rect(aes(xmin = 1.5, xmax = 2.5,
                ymin = 31, ymax = 52),
                fill = "pink", alpha = 0.02)+
  #CD8
  geom_rect(aes(xmin = 2.5, xmax = 3.5,
                ymin = 18, ymax = 35),
                fill = "pink", alpha = 0.02)+
  #B
  geom_rect(aes(xmin = 3.5, xmax = 4.5,
                ymin = 11.3, ymax = 18),
                fill = "pink", alpha = 0.02)+ 
  #MO
  geom_rect(aes(xmin = 4.5, xmax = 5.5,
                ymin = 0, ymax = 8),
                fill = "pink", alpha = 0.02)+  
  #NK
  geom_rect(aes(xmin = 5.5, xmax = 6.5,
                ymin = 4, ymax = 27),
                fill = "pink", alpha = 0.02)+  
  geom_jitter(width = 0.08, size = 3, aes(size = 3))+
  xlab("")+
  ylab("Percent Cell Type (%)")+
  scale_y_continuous(breaks = seq(0,100,10),
                     limits = c(-1,100))+ #set the y axis from 0-100 labeled every 10 units
  theme(text = element_text(size=35))+ #makes text bigger
  labs(fill = "Estimated Cell Types")+ #change the legend title
  theme(axis.text.x = element_text(colour = "black", size = 35),
        axis.text.y = element_text(colour = "black"))+ #change axis text color
  theme(panel.background = element_rect(fill = "white", #this is the background
                                colour = "black", #this is the border
                                size = 0.1, linetype = "solid"))+
  #scale_fill_manual(values=c("#F8766D", "#00BFC4", "yellow"))+
  theme(legend.position="none")+
  theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())

dev.off()
```


# Set up to graph the 3 estimate figures as a panel
```{r}
library(ggplot2)
wbc_all_samp <- pd_final_all_wbc[ , c(2, 11, 65:70)]
colnames(wbc_all_samp) #60 obs, 8 vars

#using wbc_all_samp file to start
ic_epi_ora_wbc <- wbc_all_samp[wbc_all_samp$celltype != "large", ]
table(ic_epi_ora_wbc$celltype) #4 oragene observations
# CD45pos oragene   whole 
#      20       4      18 

#turn the WBC columns wide to long
library(tidyr)
all_wbc_est_long <- gather(ic_epi_ora_wbc, wbc_cell_type, estimate, GR:NK, factor_key=TRUE)
View(all_wbc_est_long)

#turn decimal estimates into percents
all_wbc_est_long$estimate <- all_wbc_est_long$estimate*100

#fix names of GR:NK to Granulocyte:Natural Killer
table(all_wbc_est_long$wbc_cell_type) #42 of each wbc type

all_wbc_est_long$wbc_cell_type <- gsub("GR", "Granulocyte", all_wbc_est_long$wbc_cell_type)
all_wbc_est_long$wbc_cell_type <- gsub("B", "B cell", all_wbc_est_long$wbc_cell_type)
all_wbc_est_long$wbc_cell_type <- gsub("MO", "Monocyte", all_wbc_est_long$wbc_cell_type)
all_wbc_est_long$wbc_cell_type <- gsub("NK", "Natural Killer", all_wbc_est_long$wbc_cell_type)
table(all_wbc_est_long$wbc_cell_type)

#make the celltype saliva names clear
all_wbc_est_long$celltype <- gsub("CD45pos", "Immune Fractions", all_wbc_est_long$celltype)
all_wbc_est_long$celltype <- gsub("whole", "Whole Samples", all_wbc_est_long$celltype)
all_wbc_est_long$celltype <- gsub("oragene", "Oragene Kits", all_wbc_est_long$celltype)
table(all_wbc_est_long$celltype)

#make these factors in order
all_wbc_est_long$wbc_cell_type <- factor(all_wbc_est_long$wbc_cell_type,
                                             levels = c("Granulocyte",
                                                        "CD4",
                                                        "CD8",
                                                        "B cell",
                                                        "Monocyte",
                                                        "Natural Killer"))
```


# Graph the 3 estimate figures as a panel
```{r}
png("06-19-19 03-03-20 WBC prediction from Reinius - Panel.png", width=5500, height=1200)
ggplot(data = all_wbc_est_long, aes(x = wbc_cell_type, y = estimate, fill = wbc_cell_type))+
  geom_point(alpha = 0.01)+
  #GR
  geom_rect(aes(xmin = 0.5, xmax = 1.5,
                ymin = 38, ymax = 72),
                fill = "pink", alpha = 0.02)+
  #CD4
  geom_rect(aes(xmin = 1.5, xmax = 2.5,
                ymin = 31, ymax = 52),
                fill = "pink", alpha = 0.02)+
  #CD8
  geom_rect(aes(xmin = 2.5, xmax = 3.5,
                ymin = 18, ymax = 35),
                fill = "pink", alpha = 0.02)+
  #B
  geom_rect(aes(xmin = 3.5, xmax = 4.5,
                ymin = 6, ymax = 27),
                fill = "pink", alpha = 0.02)+ 
  #MO
  geom_rect(aes(xmin = 4.5, xmax = 5.5,
                ymin = 0, ymax = 8),
                fill = "pink", alpha = 0.02)+  
  #NK
  geom_rect(aes(xmin = 5.5, xmax = 6.5,
                ymin = 4, ymax = 27),
                fill = "pink", alpha = 0.02)+  
  geom_jitter(width = 0.08, size = 5, aes(size = 20))+
  xlab("")+
  ylab("Percent Cell Type (%)")+
  scale_y_continuous(breaks = seq(0,100,10),
                     limits = c(-1,100))+ #set the y axis from 0-100 labeled every 10 units
  theme(text = element_text(size=55))+ #makes text bigger
  labs(fill = "Estimated Cell Types")+ #change the legend title
  theme(axis.text.x = element_text(colour = "black", size = 55),
        axis.text.y = element_text(colour = "black", size = 55),
        axis.title.y = element_text(size = 65))+ #change axis text color
  theme(panel.background = element_rect(fill = "white", #this is the background
                                colour = "black", #this is the border
                                size = 0.1, linetype = "solid"))+
  #scale_fill_manual(values=c("#F8766D", "#00BFC4", "yellow"))+
  theme(legend.position="none")+
  theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())+
  facet_wrap(vars(celltype))+ #makes a panel
  theme(strip.background = element_blank(),
        strip.text.x = element_blank())
dev.off()
```


## 6/19/19
#_____________________________________________________________________________________________________#
# These violin plots have all saliva sample types on one plot for each WBC type
# Make violin plots of cell types - Granulocytes
```{r}
#make GR violin plots
library(ggplot2)
ggplot(pd_final_all_wbc, aes(x = celltype, y = GR))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))
```

# Make violin plots of cell types - CD4
```{r}
ggplot(pd_final_all_wbc, aes(x = celltype, y = CD4))+
  scale_y_continuous(name="CD4", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))
```

# Make violin plots of cell types - CD8
```{r}
ggplot(pd_final_all_wbc, aes(x = celltype, y = CD8))+
  scale_y_continuous(name="CD8", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))
```

# Make violin plots of cell types - B cells
```{r}
ggplot(pd_final_all_wbc, aes(x = celltype, y = B))+
  scale_y_continuous(name="B cells", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))
```

# Make violin plots of cell types - MO
```{r}
ggplot(pd_final_all_wbc, aes(x = celltype, y = MO))+
  scale_y_continuous(name="MO", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))
```

# Make violin plots of cell types - NK
```{r}
ggplot(pd_final_all_wbc, aes(x = celltype, y = NK))+
  scale_y_continuous(name="NK", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))
```

# PDF of all cell type estimation violin plots
```{r}
pdf("06-29-19 WBC type estimation.pdf")
ggplot(pd_final_all_wbc, aes(x = celltype, y = GR))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))+
  geom_rect(aes(xmin = 0, xmax = 6, ymin = 0.32, ymax = 0.64),
                   fill = "pink", alpha = 0.02)

# Make violin plots of cell types - CD4
ggplot(pd_final_all_wbc, aes(x = celltype, y = CD4))+
  scale_y_continuous(name="CD4", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))+
  geom_rect(aes(xmin = 0, xmax = 6, ymin = 0.31, ymax = 0.52),
                   fill = "pink", alpha = 0.02)

# Make violin plots of cell types - CD8
ggplot(pd_final_all_wbc, aes(x = celltype, y = CD8))+
  scale_y_continuous(name="CD8", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))+
  geom_rect(aes(xmin = 0, xmax = 6, ymin = 0.18, ymax = 0.35),
                   fill = "pink", alpha = 0.02)

# Make violin plots of cell types - B cells
ggplot(pd_final_all_wbc, aes(x = celltype, y = B))+
  scale_y_continuous(name="B cells", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))+
  geom_rect(aes(xmin = 0, xmax = 6, ymin = 0.048, ymax = 0.27),
                   fill = "pink", alpha = 0.02)

# Make violin plots of cell types - MO
ggplot(pd_final_all_wbc, aes(x = celltype, y = MO))+
  scale_y_continuous(name="MO", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))+
  geom_rect(aes(xmin = 0, xmax = 6, ymin = 0.02, ymax = 0.08),
                   fill = "pink", alpha = 0.02)

# Make violin plots of cell types - NK
ggplot(pd_final_all_wbc, aes(x = celltype, y = NK))+
  scale_y_continuous(name="NK", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))+
  geom_rect(aes(xmin = 0, xmax = 6, ymin = 0, ymax = 0.01),
                   fill = "pink", alpha = 0.02)
dev.off()
```

# PNG of all cell type estimation violin plots
```{r}
# Make violin plots of cell types - GR
png("06-29-19 WBC type estimation - GR.png")
ggplot(pd_final_all_wbc, aes(x = celltype, y = GR))+
  scale_y_continuous(name="GR", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))
dev.off()
```
```{r}
# Make violin plots of cell types - CD4
png("06-29-19 WBC type estimation - CD4.png")
ggplot(pd_final_all_wbc, aes(x = celltype, y = CD4))+
  scale_y_continuous(name="CD4", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))
dev.off()
```
```{r}
# Make violin plots of cell types - CD8
png("06-29-19 WBC type estimation - CD8.png")
ggplot(pd_final_all_wbc, aes(x = celltype, y = CD8))+
  scale_y_continuous(name="CD8", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))
dev.off()
```
```{r}
png("06-29-19 WBC type estimation - B.png")
# Make violin plots of cell types - B cells
ggplot(pd_final_all_wbc, aes(x = celltype, y = B))+
  scale_y_continuous(name="B cells", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))
dev.off()
```
```{r}
png("06-29-19 WBC type estimation - MO.png")
# Make violin plots of cell types - MO
ggplot(pd_final_all_wbc, aes(x = celltype, y = MO))+
  scale_y_continuous(name="MO", limits=c(0, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))
dev.off()
```
```{r}
png("06-29-19 WBC type estimation - NK.png")
# Make violin plots of cell types - NK
ggplot(pd_final_all_wbc, aes(x = celltype, y = NK))+
  scale_y_continuous(name="NK", limits=c(0, 1.0))+
  scale_x_discrete(name = "celltype", expand=c(.08,0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))
dev.off()
```


# Create image panel of WBC estimations
```{r}
library(ggpubr)

png("06-29-19 WBC type estimation - Panel.png", width = 800, height = 600)
# Make violin plots of cell types - GR
p1 <- ggplot(pd_final_all_wbc, aes(x = celltype, y = GR))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.title.y = element_text(size=14, face = "bold"))+
  theme(axis.title.x=element_blank())+
  geom_rect(aes(xmin = 0, xmax = 5, ymin = 0.32, ymax = 0.64),
                   fill = "pink", alpha = 0.02)

# Make violin plots of cell types - CD4
p2 <- ggplot(pd_final_all_wbc, aes(x = celltype, y = CD4))+
  scale_y_continuous(name="CD4", limits=c(-0.01, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.title.y = element_text(size=14, face = "bold"))+
  theme(axis.title.x=element_blank())+
  geom_rect(aes(xmin = 0, xmax = 5, ymin = 0.31, ymax = 0.52),
                   fill = "pink", alpha = 0.02)

# Make violin plots of cell types - CD8
p3 <- ggplot(pd_final_all_wbc, aes(x = celltype, y = CD8))+
  scale_y_continuous(name="CD8", limits=c(-0.01, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.title.y = element_text(size=14, face = "bold"))+
  theme(axis.title.x=element_blank())+
  geom_rect(aes(xmin = 0, xmax = 5, ymin = 0.18, ymax = 0.35),
                   fill = "pink", alpha = 0.02)

# Make violin plots of cell types - B cells
p4 <- ggplot(pd_final_all_wbc, aes(x = celltype, y = B))+
  scale_y_continuous(name="B cells", limits=c(-0.01, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.title.y = element_text(size=14, face = "bold"))+
  theme(axis.title.x=element_blank())+
  geom_rect(aes(xmin = 0, xmax = 5, ymin = 0.048, ymax = 0.27),
                   fill = "pink", alpha = 0.02)

# Make violin plots of cell types - MO
p5 <- ggplot(pd_final_all_wbc, aes(x = celltype, y = MO))+
  scale_y_continuous(name="MO", limits=c(-0.01, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.title.y = element_text(size=14, face = "bold"))+
  theme(axis.title.x=element_blank())+
  geom_rect(aes(xmin = 0, xmax = 5, ymin = 0.02, ymax = 0.08),
                   fill = "pink", alpha = 0.02)

# Make violin plots of cell types - NK
p6 <- ggplot(pd_final_all_wbc, aes(x = celltype, y = NK))+
  scale_y_continuous(name="NK", limits=c(-0.01, 1.0))+
  geom_violin()+
  geom_jitter(shape = 16, position = position_jitter(0.05))+
  theme(axis.text.x = element_text(size=12))+
  theme(axis.title.y = element_text(size=14, face = "bold"))+
  theme(axis.title.x=element_blank())+
  geom_rect(aes(xmin = 0, xmax = 5, ymin = 0.04, ymax = 0.27),
                   fill = "pink", alpha = 0.02)

ggarrange(p1, p2, p3, p4, p5, p6, ncol = 3, nrow = 2)
dev.off()
```



#________________________________________________________________________________________________
# ANOVA of cell types
```{r}
GR_anova <- aov(GR~celltype, data = pd_final_all_wbc)

CD4_anova <- aov(CD4~celltype, data = pd_final_all_wbc)

CD8_anova <- aov(CD8~celltype, data = pd_final_all_wbc)

B_anova <- aov(B~celltype, data = pd_final_all_wbc)

MO_anova <- aov(MO~celltype, data = pd_final_all_wbc)

NK_anova <- aov(NK~celltype, data = pd_final_all_wbc)
```

# ANOVA of cell types - Summary
```{r}
summary(GR_anova)  #F value  Pr(>F)    
                   #47.98    6.95e-12
summary(CD4_anova) #F value Pr(>F)    
                   #51.89   2.07e-12
summary(CD8_anova) #F value Pr(>F)    
                   #0.091  0.913
summary(B_anova)   #F value Pr(>F)    
                   #37.22  2.89e-10
summary(MO_anova)  #F value Pr(>F)    
                   #43.01 3.61e-11
summary(NK_anova)  #F value Pr(>F)    
                   #22.27 1.9e-07
```



#____________________________________________________________________________________________________#
#____________________________________________________________________________________________________#
# This is not that relevant because it's only the most variable probes
```{r}
#variance of every probe;
vars <- rowVars(as.matrix(beta_final_all))
summary(vars)
names(vars) <- rownames(beta_final_all)
head(vars)

#order the variance, pick out top 1000 highest variance;
vars_order <- vars[order(-vars)]
var_beta <- beta_final_all[rownames(beta_final_all) %in% names(vars_order[1:1000]),]
dim(var_beta)
```
```{r}
library(gplots)
#summary by groups aka cell types;
tapply(pd_final_all_wbc$GR, pd_final_all_wbc$celltype, summary)

#output to pdf
pdf('06-19-19 Heatmaps of WBCs.pdf')

#heatmap based on most WBC fractions - Granulocyte (split by median)
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final_all_wbc$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final_all_wbc$GR>0.73)],
          main="Top Variable Probes - Granulocytes")
#more hypermethylated 45+ are granulocytes;

#heatmap based on most WBC fractions - CD4
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final_all_wbc$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final_all_wbc$CD4>0.44)],
          main="Top Variable Probes - CD4 (CD45pos median)")
#explains some of the hypermethylation in the large cells?

#heatmap based on most WBC fractions - CD4
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final_all_wbc$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final_all_wbc$CD4>0.17)],
          main="Top Variable Probes - CD4 (large cell median)")

#heatmap based on most WBC fractions - B
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final_all_wbc$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final_all_wbc$B>0.18)],
          main="Top Variable Probes - B cells")
#large has some B cells (blue)

#heatmap based on most WBC fractions - Monocyte
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final_all_wbc$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final_all_wbc$MO>0.09)],
          main="Top Variable Probes - Monocytes")
#large has some monocytes (blue)

#heatmap based on most WBC fractions - CD8
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final_all_wbc$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final_all_wbc$CD8>0.0034)],
          main="Top Variable Probes - CD8")
#almost nothing

#heatmap based on most WBC fractions - NK
heatmap.2(as.matrix(var_beta), labRow=NA, col=redblue(256),
          density.info="none", trace="none", 
          dendrogram="column", labCol=pd_final_all_wbc$Sample.ID,
          ColSideColors=myColors[as.factor(pd_final_all_wbc$NK>0.004)],
          main="Top Variable Probes - NK")
dev.off()
```